---
title: "Locus and sample QC and genotype counts"
author: "Jelmer Poelstra"
date: today
format:
  html:
    self-contained: true
    cache: false
    number-sections: true
    number-depth: 2
    toc: true
    toc-depth: 3
    fig-dpi: 300
    df-print: kable
    grid:
      body-width: 800px
knitr:
  opts_chunk:
    message: FALSE
    out.width: "85%"
    R.options:
      knitr.table.html.attr: "quarto-disable-processing=true"
execute:
  echo: false
  warning: false
editor_options:
  chunk_output_type: console
---

-------

<br>

```{r}
# --- Setup
# Load packages
library(tidyverse)
library(gt)
library(ggtext)
library(ggiraph)
library(here)

# Settings
theme_set(theme_classic(base_size = 14))
theme_update(plot.title = element_text(face = "bold", hjust = 0.5),
             plot.title.position = "plot")

# ggplot picks a plot's default scale by looking up `scale_<aes>_<type>` in the
# global environment before its own namespace, so shadowing it here doubles the
# x-axis break density everywhere at once. Plots that pass their own `breaks`
# still win, since `breaks` takes precedence over `n.breaks`.
scale_x_continuous <- function(...) ggplot2::scale_x_continuous(n.breaks = 10, ...)

# --- Table helpers
# Columns whose values are all whole numbers are printed without decimals;
# everything else numeric gets `digits` decimals, so no table shows 5+ decimals
is_whole <- function(x) is.numeric(x) && all(x == round(x), na.rm = TRUE)

# Tables are left at their natural width unless `full_width` is set, since most
# of them have a handful of narrow columns and look better not stretched out
qc_gt <- function(df, digits = 2, title = NULL, full_width = FALSE) {
  tab <- df |>
    gt() |>
    fmt_number(
      columns = where(\(x) is.numeric(x) && !is_whole(x)),
      decimals = digits
    ) |>
    fmt_integer(columns = where(is_whole)) |>
    sub_missing(missing_text = "–") |>
    cols_label_with(fn = \(x) str_replace_all(x, "_", " ")) |>
    tab_options(
      table.width = if (full_width) pct(100) else "auto",
      table.font.size = px(15),
      data_row.padding = px(3),
      column_labels.font.weight = "bold",
      heading.title.font.weight = "bold"
    )
  if (!is.null(title)) tab <- tab |> tab_header(title = title)
  tab
}

# For the descriptive pipeline tables, whose cells contain inline code
doc_gt <- function(df, full_width = TRUE) {
  df |>
    gt() |>
    fmt_markdown(columns = where(is.character)) |>
    sub_missing(missing_text = "–") |>
    cols_label_with(fn = \(x) str_replace_all(x, "_", " ")) |>
    tab_options(
      table.width = if (full_width) pct(100) else "auto",
      table.font.size = px(15),
      data_row.padding = px(3),
      column_labels.font.weight = "bold"
    )
}

# Define input files (all written by scripts/05a_prep-qc-tables.R)
indir <- here("results/stats/summaries")
all_file <- file.path(indir, "sample-by-locus-stats.tsv")
locus_file <- file.path(indir, "per-locus-stats.tsv")
sample_file <- file.path(indir, "per-sample-stats.tsv")
summary_file <- file.path(indir, "filtering-summary.tsv")

# Written by 04g, not 05a: depth against distance from the capture-probe edge
profile_file <- here("results/stats/depth_profile/depth_profile.tsv")

# Written by the 4I side runs of 04b and 04f, which scored the flagged samples
# together with the retained ones
sex_flagged_file <- here("results/stats/sex_z_flagged/sex_calls.tsv")
vcf_stats_flagged_file <- here("results/stats/vcf_tstv_flagged/per_sample_vcf_stats.tsv")

# Written by 04h: every het genotype that reached the 02b allele-balance filter
ab_files <- list.files(here("results/stats/allele_balance"),
                       pattern = "_het-ab\\.tsv\\.gz$", full.names = TRUE)

# Read input files. `z_class` is NA for every autosomal locus, which is the
# large majority, so readr guesses logical from the leading rows and silently
# turns each real class into NA unless the type is given.
all <- read_tsv(all_file, col_types = cols(z_class = "c", .default = col_guess()))
per_locus <- read_tsv(locus_file, show_col_types = FALSE)
per_sample <- read_tsv(sample_file, show_col_types = FALSE)
filtering_summary <- read_tsv(summary_file, show_col_types = FALSE)
dp_profile <- read_tsv(profile_file, show_col_types = FALSE)
sex_flagged <- read_tsv(sex_flagged_file, show_col_types = FALSE)
vcf_stats_flagged <- read_tsv(vcf_stats_flagged_file, show_col_types = FALSE)
# `ao` is '.' where freebayes reports no alternate observation
het_ab <- read_tsv(ab_files, id = "sample", na = c("NA", "."),
                   col_types = "cicdddnc") |>
  mutate(sample = str_remove(basename(sample), "_het-ab\\.tsv\\.gz$"))

# The sample-by-locus table still holds the loci dropped by 04b/04c, so that the
# sections below can compare them against the retained ones; everything that is
# meant to describe the final data set filters on `retained_final` first
all_final <- all |> filter(retained_final)
locus_final <- per_locus |> filter(retained_final)
sample_final <- per_sample |> filter(retained)

# All three genotype metrics per sample, used from the I20937 section onward
geno_smp <- all_final |>
  filter(genotype != "n_miss") |>
  # Mean of per-locus percentages; 05a's `het_pct_final` pools sites across loci
  # instead, so it weights long loci more. NA where a locus is uncalled in this
  # sample
  summarize(geno_pct = mean(geno_pct, na.rm = TRUE), .by = c(sample, genotype)) |>
  left_join(sample_final, by = "sample")

het_per_sample <- geno_smp |>
  filter(genotype == "n_het") |>
  rename(pct_het = geno_pct)

n_contam <- sum(per_sample$contam, na.rm = TRUE)

# Samples with a median allele balance below this are listed in the
# flagged-samples section and counted in the summary
ab_low <- 0.45
n_low_ab_flagged <- sum(per_sample$contam %in% TRUE & per_sample$ab_median < ab_low, na.rm = TRUE)
n_low_ab_retained <- sum(sample_final$ab_median < ab_low, na.rm = TRUE)
```

## Summary

- `r nrow(sample_final)` of `r nrow(per_sample)` samples and
  `r nrow(locus_final)` of `r nrow(per_locus)` loci are retained.
- There are two sequencing batches, and they differ:
  capture worked about twice as well in the 605 run than in the other run.
- Short (~200 bp) baits have about half the depth of long ones because of capture
  geometry, but in the end, only modest differences in usable data arise between
  locus types --- because depth is quite high overall.
- Paralog filters: `r sum(per_locus$stage == "dropped: excess het")` loci are
  dropped for excess heterozygosity and
  `r sum(per_locus$stage == "dropped: allele balance")` on allele balance alone;
  all `r sum(str_starts(per_locus$stage, "dropped: Z"))` sex-linked loci are
  dropped.
- Four samples (`I1495`, `I1507`, `I1624` and `I1517`) not flagged as contaminants
  by Felipe look contaminated, and are removed along with the flagged ones
  (see *Samples to watch*).
- One sample (`I20937`) rather stands out by singleton count,
  but this is likely due to species identity: it is the only *B. jabrensis*.
- Genotype counts by species and by *jararaca* state of origin are in the last section.

## Pipeline overview

### Step 1 --- Locus definition by BLAST

Probes are BLASTed against the reference genome and the hits are filtered:

```{r}
tribble(
  ~Step, ~Setting, ~Value,
  "BLAST", "`--top_n_query`", "5 hits per probe",
  "BLAST", "`--pct_id`", "≥ 90%",
  "BLAST", "`--pct_qcov` (HSP-based)", "≥ 80%",
  "`01d_filter-loci.R`", "Percent identity (`pident`)", "≥ 95% for at least one hit",
  "`01d_filter-loci.R`", "Query coverage (`qcovhsp`)", "≥ 90% for at least one hit",
  "`01d_filter-loci.R`", "Multi-hit probes", "Removed entirely (probes with > 1 BLAST hit)",
  "BED output", "Buffer around each hit", "300 bp on each side (`buffer_size`)"
) |>
  doc_gt()
```

Result: 2,000 probes → 1,936 with BLAST hits → 1,819 loci.

### Step 2 --- Genotype calling and consensus generation

A) Read processing and variant calling ---
   uses the nf-core Sarek pipeline, run per sample:

```{r}
tribble(
  ~Step, ~Tool, ~Setting,
  "Trimming", "fastp 0.24.0", "Adapter trimming, remove reads <15 bp, otherwise fastp defaults",
  "Mapping", "bwa-mem 0.7.18", "Default settings",
  "Duplicates", "GATK4 MarkDuplicates", "Duplicates marked — these are then ignored by freebayes",
  "Base recalibration", "—", "Skipped (`skip_tools: baserecalibrator`)",
  "Calling regions", "—", "Restricted to the probe BED (`intervals`)",
  "Variant calling", "freebayes 1.3.10", "`--report-monomorphic --genotype-qualities`, otherwise default"
) |>
  doc_gt()
```

B) Genotype filtering and consensus generation
   (eight `bcftools` steps):

```{r}
tribble(
  ~Step, ~Setting, ~Value, ~Effect,
  "0\\. Block splitting", "—", "—", "Multi-bp monomorphic records split into one per base",
  "1\\. Genotype quality", "`--filter_expr`", "`FORMAT/DP < 5 || FORMAT/GQ < 20`", "Genotypes below depth/GQ set to missing (`./.`)",
  "2\\. Site quality", "`--min_qual`", "30", "Genotypes at low-QUAL variant sites set to missing",
  "3\\. Indel proximity", "`--snp_gap`", "5 bp", "Genotypes at SNPs within 5 bp of an indel set to missing",
  "4\\. Normalization", "`bcftools norm`", "`--multiallelics -any --atomize`", "Splits multiallelics and MNVs, left-aligns indels",
  "5\\. Indels", "`bcftools view`", "`-V indels`", "Indels removed; SNPs only",
  "6\\. Allele balance", "`--min_ab` / `--max_ab`", "0.25 / 0.75", "Skewed heterozygote genotypes set to missing (only where AO + RO is within 2 reads of DP, `--ab_tol`)",
  "7\\. Multiallelics", "`--drop_multiallelic`", "`true`", "Positions left with >1 record after step 4 are removed"
) |>
  doc_gt()
```

_(Step 0 exists because freebayes occasionally merges a run of invariant positions_
_into a single record (REF several bp long, ALT `.`), which can span past the end_
_of the per-region reference slice given to `bcftools consensus` and crash it.)_

Consensus sequences are then generated with `bcftools consensus`, where
heterozygous sites get IUPAC ambiguity codes and missing/absent genotypes
become `N`.

### Step 3 --- Pivot, trim, and filter consensus sequences

A) Pivot to locus-centric FASTA files.

B) Per-locus trimming with ClipKIT to remove the terminal `N` stretches caused
   by the buffers (terminal column is trimmed if >25% of sequences have `N` at that position).

C) Sample and locus filtering --- `03c_filter-loci.sh`, applied in this order:

```{r}
tribble(
  ~Step, ~Parameter, ~Value, ~Effect,
  "1\\. A priori sample removal", "`--remove_samples`", "—", "Samples flagged as contaminated in `metadata/metadata-final.tsv`, plus four that look contaminated in this report (`metadata/contam_samples_qc.txt`)",
  "2\\. Per-sample missingness", "`max_missing_sample`", "0.1", "Samples with mean >10% `N` across all loci removed",
  "3\\. Minimum locus length", "`min_locus_length`", "150 bp", "Post-trimming alignment length",
  "4\\. Per-locus missingness", "`max_missing_locus`", "0.2", "Loci with mean >20% `N` across retained samples removed"
) |>
  doc_gt()
```

### Step 4 --- Sex-linked loci

Sample sex is called from the Z:autosome depth ratio, and each Z locus is
classified as _hemizygous_, _PAR-like_, _gametolog_, or _paralog_, using 
a per-site excess-heterozygosity statistic.

- _PAR-like_: locus in a pseudoautosomal-like region,
  present in two copies in both sexes and behaving like an autosomal locus.
- _Hemizygous_: single-copy Z locus present only once in the heterogametic sex,
  so it shows no heterozygosity there and reduced depth relative to autosomes.
- _Gametolog_: locus with a diverged W-linked (or Y-linked) counterpart,
  so reads from both copies collapse onto the Z reference and appear as excess
  heterozygosity in the heterogametic sex.
- _Paralog_: locus duplicated elsewhere in the genome (not sex-linked divergence),
  producing excess heterozygosity from collapsed paralogous copies in both sexes.

Nevertheless, **every** sex-linked locus is currently dropped.

### Step 5 --- Paralog filter: excess heterozygosity and allele balance

Paralogous loci are inferred using two tests, and a locus failing either test is dropped:

- The first test uses a per-site statistic `f`, the fraction of called samples that
  are heterozygous.
- The second uses allele balance: where one of two collapsed copies carries a variant, a
  het shows it on a quarter to a third of the reads rather than half, in most
  samples that are heterozygous there.

```{r}
tribble(
  ~Parameter, ~Value, ~Effect,
  "`min_call_frac`", "0.25", "A site is only scored when at least this fraction of samples has a called genotype",
  "`max_het_frac`", "0.6", "A scored site with `f` above this is an 'excess-het site' (the HWE maximum is 0.5)",
  "`max_site_frac`", "0.005", "A locus is dropped when more than this fraction of its scored sites are excess-het sites...",
  "`min_het_sites`", "3", "...and it has at least this many excess-het sites",
  "`min_ab_hets`", "5", "A sample counts toward a locus's allele-balance test when it has at least this many evaluated hets there...",
  "`ab_skew_cutoff`", "0.4", "...and the locus is skewed in that sample when their median allele balance is below this",
  "`min_ab_samples`", "20", "A locus is only judged on allele balance with at least this many counted samples",
  "`max_ab_skew_frac`", "0.5", "A locus is dropped when it is skewed in more than this fraction of counted samples"
) |>
  doc_gt()
```

## Filtering summary

Loci and samples remaining after each filtering step:

```{r}
filtering_summary |> qc_gt(full_width = TRUE)
```

Step at which each locus and sample was dropped:

```{r}
per_locus |> count(stage, name = "n_loci") |> qc_gt()
per_sample |> count(stage, name = "n_samples") |> qc_gt()
```

## Depth of coverage

### Per sample

The depth figures in this section are mean FORMAT/DP over called genotypes
within the retained locus intervals.
These are a capture probe intervals plus what remains of a 300-bp (600 bp total)
buffer on each side after ClipKIT trimming of poorly covered ends
(a median of ~350 bp of buffer remains on short loci and ~310 bp on long ones).
Because the buffer is much less deeply covered than the probe,
loci of different lengths are not directly comparable.

```{r}
sample_final |>
  ggplot(aes(x = mean_dp)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black") +
  labs(
    title = "Mean depth of coverage per sample",
    x = "Mean depth of coverage",
    y = "Frequency (samples)"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))
```

#### The depth outlier: I1767

I1767 has by far the highest depth (bar on far right in plot above).
But it looks like it was simply sequenced deeper,
as other stats for this sample are not unusual.
Duplication at 10.8% is the second-highest, but that makes sense given the depth.

```{r}
sample_final |>
  select(sample, species, mean_dp, n_reads, pct_on_target, enrichment,
         pct_duplication, error_rate, mean_pct_N_final, ts_tv, n_singletons) |>
  arrange(desc(mean_dp)) |>
  head(5) |>
  qc_gt(digits = 4)
```

### Per locus

```{r}
locus_final |>
  ggplot(aes(x = mean_dp)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "black") +
  labs(
    title = "Mean depth of coverage per locus",
    x = "Mean depth of coverage (averaged over samples)",
    y = "Frequency (loci)"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

# Depth is the main driver of missingness, so these should be related
locus_final |>
  ggplot(aes(x = mean_dp, y = 100 * (1 - mean_call_frac))) +
  geom_point(alpha = 0.3, size = 1.5, color = "steelblue") +
  labs(
    title = "Locus depth vs. locus missingness",
    x = "Mean depth of coverage",
    y = "Mean percentage of missing data (N)"
  )
```

## Mapping and capture

Per-sample stats:

```{r}
sample_final |>
  select(sample, n_reads, pct_mapped, pct_properly_paired, pct_mq0,
         pct_duplication, pct_on_target, target_mean_dp, enrichment) |>
  summarize(across(!sample, list(min = min, median = median, max = max))) |>
  pivot_longer(
    everything(),
    names_to = c("metric", "stat"),
    names_pattern = "(.*)_(min|median|max)"
  ) |>
  pivot_wider(names_from = stat, values_from = value) |>
  gt() |>
  # Read counts are in the millions and the rates are percentages, so a shared
  # format would render one of them unreadably
  fmt_number(columns = c(min, median, max), decimals = 2) |>
  fmt_integer(columns = c(min, median, max), rows = metric == "n_reads") |>
  cols_label_with(fn = \(x) str_replace_all(x, "_", " ")) |>
  tab_options(
    table.width = pct(100),
    table.font.size = px(15),
    data_row.padding = px(3),
    column_labels.font.weight = "bold"
  )
```

```{r}
# The share of all sequenced bases that landed inside the probe intervals. The
# targets are 2.2 Mb of a 1.57 Gb genome, so even a modest percentage here is a
# large enrichment. Filled by batch because the two runs separate almost
# completely on this axis -- see the section below.
sample_final |>
  ggplot(aes(x = pct_on_target, fill = seq_batch)) +
  geom_histogram(binwidth = 0.25, color = "black") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "On-target rate per sample",
    x = "Percentage of sequenced bases inside the probe intervals",
    y = "Frequency (samples)",
    fill = "Batch"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

# Read count sets the ceiling on depth, but the conversion rate differs per
# batch, so there are two trends rather than one. Coloring by duplicate rate
# here would be misleading: duplication is itself higher in the 605 batch, so
# the gradient would mostly re-encode the batch.
sample_final |>
  ggplot(aes(x = n_reads / 1e6, y = mean_dp, color = seq_batch)) +
  geom_point(size = 2.5, alpha = 0.8) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 0.7) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "Read count vs. depth",
    x = "Read count (millions)",
    y = "Mean depth of coverage",
    color = "Batch"
  )
```

### Sequencing batch

The samples were sequenced in two runs, and the capture (in terms of the on-target rate)
worked about twice as well in one of them ---
hence the divergence in the two plots above.

```{r}
sample_final |>
  summarize(
    n_samples = n(),
    across(c(n_reads, pct_on_target, enrichment, pct_duplication, mean_dp,
             mean_pct_N_final),
           mean),
    dp_per_Mread = mean(mean_dp / (n_reads / 1e6)),
    .by = seq_batch
  ) |>
  arrange(seq_batch) |>
  relocate(dp_per_Mread, .after = mean_dp) |>
  qc_gt(title = "Per-batch means", full_width = TRUE)

# Loci called per sample, per batch, for the text below
loci_called <- sample_final |>
  summarize(mean = round(mean(n_loci_called)), min = min(n_loci_called),
            .by = seq_batch)
lc <- function(b, col) loci_called[[col]][loci_called$seq_batch == b]
```

This difference between runs is partly offset because the weaker-capture batch
was sequenced about 30% deeper and it has a lower duplicate rate.
The net effect on the final data is modest:

- Mean missingness after filtering is
  `r round(mean(sample_final$mean_pct_N_final[sample_final$seq_batch == "605_HCTGLBCXX"]), 2)`%
  in the 605 batch versus
  `r round(mean(sample_final$mean_pct_N_final[sample_final$seq_batch == "705_H2HLLBCXY"]), 2)`%
  in the 705 batch
- The number of loci called per sample is essentially identical
  (mean `r lc("605_HCTGLBCXX", "mean")` and `r lc("705_H2HLLBCXY", "mean")` of `r nrow(locus_final)`).
- Of the `r sum(per_sample$stage == "dropped: missingness")` samples dropped due
  to high missingness,
  `r sum(per_sample$stage == "dropped: missingness" & per_sample$seq_batch == "705_H2HLLBCXY", na.rm = TRUE)`
  are from the 705 batch.

Batch is not confounded with the main biological groupings: the split is close
to even for *jararaca* and for the mainland samples.
However, it is not even for the island endemics:
*insularis* and *sazimai* were sequenced only in the 605 batch and most *alcatraz*
samples in the 705 batch.

```{r}
sample_final |>
  count(land_type, species, seq_batch) |>
  pivot_wider(names_from = seq_batch, values_from = n, values_fill = 0) |>
  arrange(land_type, species) |>
  qc_gt(title = "Samples per land type, species and batch")
```

## Locus lengths

```{r}
# All loci with a length from 03c, so the `min_locus_length` cutoff is visible
per_locus |>
  mutate(set = ifelse(retained_final, "retained", "dropped")) |>
  ggplot(aes(x = length, fill = set)) +
  geom_histogram(binwidth = 50, color = "black") +
  geom_vline(xintercept = 150, linetype = "dashed") +
  labs(x = "Locus length (bp)", y = "Frequency (loci)", fill = NULL) +
  scale_fill_manual(values = c(retained = "steelblue", dropped = "grey70")) +
  # 125-bp breaks collide on an axis this wide, so this plot keeps 250
  scale_x_continuous(breaks = seq(0, 5000, by = 250)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))
```

Locus length and depth are correlated
(r = `r round(cor(locus_final$length, locus_final$mean_dp, use = "complete.obs"), 2)`),
for two reasons:

1. **Capture "geometry".**
   Capture efficiency ramps up over the first ~300 bp from its edge,
   so longer baits will have higher depth on average:

```{r}
# Probe length class per locus, also used in the allele-balance section
probe_dp <- locus_final |>
  mutate(
    probe_len = n_sites_pre - 600,
    probe_class = ifelse(probe_len < 600, "short", "long")
  )

# The outermost bins on each side are thin -- few probes are long enough to fill
# the innermost bin of their class, and few loci sit far enough from a neighbor
# to fill the outermost buffer bin -- and they run against the trend, so bins
# holding under 10% of their class's typical bin are dropped rather than plotted
dp_profile_plot <- dp_profile |>
  filter(n_pos >= 0.1 * median(n_pos), .by = probe_class)

dp_profile_plot |>
  filter(dist_bin >= -300, dist_bin <= 600) |>
  ggplot(aes(x = dist_bin + 25, y = mean_dp, color = probe_class)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  labs(
    title = "Depth against distance from the capture-probe edge",
    subtitle = "Negative distances are buffer; 0 is the first on-probe base",
    x = "Distance from nearest probe edge (bp)",
    y = "Mean depth of coverage",
    color = "Probe class"
  ) +
  scale_x_continuous(breaks = seq(-300, 600, by = 100)) +
  scale_y_continuous(n.breaks = 8)
```

2. **Trimming.** Poorly covered loci are trimmed more severely:

```{r}
per_locus |>
  filter(retained_final) |>
  ggplot(aes(x = length, y = mean_dp, color = pct_trimmed)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "Locus length vs. depth",
    x = "Locus length after trimming (bp)",
    y = "Mean depth of coverage",
    color = "% trimmed"
  ) +
  scale_y_continuous(n.breaks = 8) +
  scale_color_viridis_c(option = "plasma", end = 0.9)
```

```{r}
# Short vs. long final loci, for the text below
by_len <- locus_final |>
  mutate(size = ifelse(length < 1000, "short", "long")) |>
  summarize(
    n = n(),
    dp = median(mean_dp),
    call = median(mean_call_frac),
    worst_call = median(min_call_frac),
    prop_var = median(prop_variable),
    n_pi = median(n_pi),
    .by = size
  )
bl <- function(size, col) by_len[[col]][by_len$size == size]
```

## End trimming

ClipKIT removed terminal based that were mostly N.
For reference: the buffer is at most 75% of an interval
(600 out of 800 bp for a short locus),
so a locus that was trimmed by 80-99% also lost part of the probe region.

Loci where ClipKIT failed because missing data was high across the whole locus
were left untrimmed and dropped at a later stage due to high missingness.

```{r}
per_locus |>
  mutate(set = case_when(
    trim_failed ~ "ClipKIT failed",
    retained_final ~ "retained",
    .default = "dropped"
  )) |>
  ggplot(aes(x = pct_trimmed, fill = set)) +
  geom_histogram(binwidth = 2, color = "black") +
  labs(
    title = "Percentage of each alignment trimmed",
    x = "Percentage of alignment columns trimmed",
    y = "Frequency (loci)",
    fill = NULL
  ) +
  scale_fill_manual(values = c(retained = "steelblue", dropped = "grey70",
                               `ClipKIT failed` = "firebrick")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

# Heavy trimming should track with low retained depth, since both follow from a
# locus that captured poorly
locus_final |>
  ggplot(aes(x = pct_trimmed, y = mean_dp)) +
  geom_point(alpha = 0.3, size = 1.5, color = "steelblue") +
  labs(
    title = "Trimming vs. depth",
    x = "Percentage of alignment columns trimmed",
    y = "Mean depth of coverage"
  )
```

## Missingness

```{r}
# `pct_N` is 03c's value over all loci before the locus filters, so the samples
# it removed show up against the `max_missing_sample` cutoff
per_sample |>
  filter(!is.na(pct_N)) |>
  mutate(set = ifelse(retained, "retained", "dropped")) |>
  ggplot(aes(x = pct_N, fill = set)) +
  geom_histogram(binwidth = 1, color = "black", linewidth = 0.2) +
  geom_vline(xintercept = 10, linetype = "dashed", color = "grey30") +
  labs(
    title = "Mean missingness per sample, before filtering",
    x = "Mean percentage of missing data (N)",
    y = "Frequency (samples)",
    fill = NULL
  ) +
  scale_fill_manual(values = c(retained = "steelblue", dropped = "grey70")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

# Depth is the main driver here too
sample_final |>
  ggplot(aes(x = mean_dp, y = mean_pct_N_final)) +
  geom_point(alpha = 0.6, size = 2.5, color = "steelblue") +
  labs(
    title = "Sample depth vs. sample missingness",
    x = "Mean depth of coverage",
    y = "Mean percentage of missing data (N)"
  )
```

## Sample sex

Snakes are ZW, so females are hemizygous over the differentiated part of the Z
and show about half the male depth there.

```{r}
sample_final |>
  mutate(sex_confident = factor(sex_confident, levels = c(TRUE, FALSE), labels = c("Yes", "No"))) |>
  ggplot(aes(x = z_depth_ratio, fill = sex, color = sex_confident)) +
  geom_histogram(binwidth = 0.01, linewidth = 0.5) +
  labs(
    title = "Z:autosome depth ratio",
    x = "Z:autosome depth ratio",
    y = "Frequency (samples)",
    fill = "Sex",
    color = "Sex confident"
  ) +
  scale_fill_brewer(palette = "Dark2") +
  scale_color_manual(values = c(Yes = "white", No = "black")) +
  # A grey key fill, so that the white outline for "Yes" is visible in the legend
  guides(color = guide_legend(override.aes = list(fill = "grey60"))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

sample_final |>
  count(sex, sex_confident) |>
  mutate(sex_confident = ifelse(sex_confident, "confident", "not confident")) |>
  pivot_wider(names_from = sex_confident, values_from = n, values_fill = 0) |>
  qc_gt()
```

Sex was tentatively assigned using the Z depth ratio,
heterozygosity in hemizygous loci,
and the number of heterozygous gametolog sites.
Samples whose sex could not be called with confidence:

```{r}
sample_final |>
  filter(!sex_confident) |>
  select(sample, species, mean_dp, z_depth_ratio, z_het_density,
         n_het_gametolog, sex, sex_flags) |>
  arrange(z_depth_ratio) |>
  qc_gt(digits = 4)
```

The one not-confident male, `I12279`, was flagged for het_density(z=3.1):
its heterozygosity density over the hemizygous Z (0.000333) is an outlier relative
to other males, even though its depth ratio (0.880) and gametolog-het count (5) both
look like a normal male.

## Excess heterozygosity and allele balance

These are the tests done to detect paralogous loci.

### Excess heterozygosity

A locus was dropped as a likely collapsed paralog when more than 0.5% of its
scored sites (and at least 3 sites) were heterozygous in more than 60% of called
samples (vs. a maximum of 50% under HWE).

```{r}
per_locus |>
  filter(!is.na(excess_site_frac)) |>
  mutate(set = ifelse(retained_final, "retained", "dropped")) |>
  ggplot(aes(x = 100 * excess_site_frac, fill = set)) +
  geom_histogram(binwidth = 0.05, color = "black") +
  # The 0.5% cutoff. It is not on its own a boundary, since dropping also needs
  # >= 3 excess-het sites, but no retained locus happens to sit right of it
  geom_vline(xintercept = 0.5, linetype = "dashed") +
  labs(
    title = "Excess-heterozygosity sites per locus",
    x = "Percentage of scored sites heterozygous in >60% of samples",
    y = "Frequency (loci)",
    fill = NULL
  ) +
  scale_fill_manual(values = c(retained = "steelblue", dropped = "grey70")) +
  # Default breaks are spaced for a linear axis and leave the log1p scale with
  # almost none, so they are given explicitly
  scale_y_continuous(
    transform = "log1p",
    breaks = c(0, 1, 3, 10, 30, 100, 300, 1000),
    expand = expansion(mult = c(0, 0.1))
  )
```

The loci dropped for excess heterozygosity, with the ones just below the
threshold for comparison:

```{r}
per_locus |>
  filter(!is.na(n_excess_sites)) |>
  mutate(
    pct_excess_sites = 100 * excess_site_frac,
    pct_max_site_het = 100 * max_site_het_frac,
    pct_het_density = 100 * het_density
  ) |>
  select(locus, stage, length, n_sites_het_scored, n_excess_sites,
         pct_excess_sites, pct_max_site_het, pct_het_density) |>
  arrange(desc(n_excess_sites)) |>
  head(15) |>
  qc_gt()
```

### Allele balance

For each sample with at least 5 heterozygous sites at a locus,
a locus was considered 'skewed' in that sample if the median balance was below 0.4.
A collapsed paralog should be skewed in most samples that are heterozygous there,
but a locus with a few noisy samples should not.

```{r}
ab_all <- het_ab |>
  filter(sample %in% sample_final$sample) |>
  left_join(select(per_locus, locus, stage, retained_final), by = "locus")

ab_final <- ab_all |> filter(retained_final)

min_ab_hets <- 5
min_ab_samples <- 20
ab_skew_cutoff <- 0.4
max_ab_skew_frac <- 0.5

ab_locus <- ab_all |>
  filter(status != "exempt") |>
  summarize(n_het = n(), median_ab = median(ab), .by = c(locus, stage, sample)) |>
  filter(n_het >= min_ab_hets) |>
  summarize(
    n_samples = n(),
    n_skewed = sum(median_ab < ab_skew_cutoff),
    pct_skewed = 100 * n_skewed / n_samples,
    .by = c(locus, stage)
  ) |>
  filter(n_samples >= min_ab_samples) |>
  left_join(
    ab_all |>
      filter(status != "exempt") |>
      summarize(n_het = n(), median_ab = median(ab),
                pct_blanked = 100 * mean(status == "blanked"), .by = locus),
    by = "locus"
  ) |>
  left_join(select(per_locus, locus, mean_dp, n_excess_sites), by = "locus")

# Pull one value for one locus, for the text
ab_l <- function(l, col) ab_locus[[col]][ab_locus$locus == l]

ab_locus |>
  summarize(
    n_loci = n(),
    median_pct_skewed = median(pct_skewed),
    max_pct_skewed = max(pct_skewed),
    .by = stage
  ) |>
  arrange(desc(median_pct_skewed)) |>
  qc_gt(title = paste0("Share of samples in which a locus is skewed, by locus fate (loci with >= ",
                       min_ab_samples, " counted samples)"))

ab_locus |>
  mutate(set = case_when(
    stage %in% c("retained", "dropped: allele balance") ~ stage,
    .default = "dropped: other"
  )) |>
  ggplot(aes(x = pct_skewed, fill = set)) +
  geom_histogram(binwidth = 2.5, boundary = 0, color = "black") +
  # The 04c cutoff (`max_ab_skew_frac`)
  geom_vline(xintercept = 100 * max_ab_skew_frac, linetype = "dashed") +
  labs(
    title = "Share of samples in which a locus has skewed allele balance",
    subtitle = paste0("Considered skewed if median balance <", ab_skew_cutoff),
    x = "Percentage of samples skewed",
    y = "Frequency (loci)",
    fill = NULL
  ) +
  scale_fill_manual(values = c(
    retained = "steelblue",
    `dropped: allele balance` = "firebrick",
    `dropped: other` = "grey70"
  )) +
  scale_y_continuous(
    transform = "log1p",
    breaks = c(0, 1, 3, 10, 30, 100, 300),
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 14))

#ab_locus |>
#  arrange(desc(pct_skewed)) |>
#  head(15) |>
#  qc_gt(title = "Loci skewed in the largest share of samples")
```

**L1826 is dropped by the allele-balance test alone.** 

## Locus informativeness

```{r}
locus_final |>
  summarize(
    across(c(n_sites_scored, n_variable, n_pi, n_singleton), sum)
  ) |>
  mutate(
    pct_variable = 100 * n_variable / n_sites_scored,
    pct_pi = 100 * n_pi / n_sites_scored
  ) |>
  qc_gt()
```

```{r}
locus_final |>
  ggplot(aes(x = n_pi)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black") +
  labs(
    title = "Parsimony-informative sites per locus",
    x = "Number of parsimony-informative sites",
    y = "Frequency (loci)"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

locus_final |>
  ggplot(aes(x = 100 * prop_pi)) +
  geom_histogram(binwidth = 0.25, fill = "steelblue", color = "black") +
  labs(
    title = "Density of parsimony-informative sites per locus",
    x = "Percentage of scored sites that are parsimony-informative",
    y = "Frequency (loci)"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

# Longer loci carry more informative sites simply by being longer; loci far below
# the trend are conserved rather than badly sequenced
locus_final |>
  ggplot(aes(x = length, y = n_pi, color = mean_dp)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_viridis_c() +
  labs(
    title = "Locus length vs. informative sites",
    x = "Locus length (bp)",
    y = "Number of parsimony-informative sites",
    color = "Depth"
  )
```

## Variant quality per sample

Real SNPs have a transition:transversion ratio of ~2, whereas random calls have 0.5.

```{r}
sample_final |>
  ggplot(aes(x = ts_tv)) +
  geom_histogram(binwidth = 0.01, fill = "steelblue", color = "black") +
  geom_vline(xintercept = 2, linetype = "dashed", color = "grey40") +
  labs(
    title = "Ts/Tv ratio per sample",
    x = "Transitions / transversions",
    y = "Frequency (samples)"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

# A ratio that falls away at low depth is the signature of depth-driven noise
sample_final |>
  ggplot(aes(x = mean_dp, y = ts_tv)) +
  geom_point(alpha = 0.7, size = 2.5, color = "steelblue") +
  geom_hline(yintercept = 2, linetype = "dashed", color = "grey40") +
  labs(
    title = "Depth vs. Ts/Tv",
    x = "Mean depth of coverage",
    y = "Transitions / transversions"
  )
```

### The singleton outlier: I20937

```{r}
sample_final |>
  ggplot(aes(x = n_singletons)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black") +
  labs(
    title = "Singleton (private) variants per sample",
    x = "Number of singletons",
    y = "Frequency (samples)"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))
```

```{r}
# Values for the text below and in 'Samples to watch'
# From `per_sample`, since the samples removed in 03c are not in `sample_final`
sv <- function(s, col) per_sample[[col]][per_sample$sample == s]
b705 <- filter(sample_final, seq_batch == "705_H2HLLBCXY")
```

I20937 has `r format(sv("I20937", "n_singletons"), big.mark = ",")` singletons
against a median of `r median(sample_final$n_singletons)`.
It is the only *B. jabrensis* in the dataset.

It does not look contaminated: heterozygosity is among the lowest in the panel
(`r sprintf("%.2f%%", het_per_sample$pct_het[het_per_sample$sample == "I20937"])`
vs a median of `r sprintf("%.2f%%", median(het_per_sample$pct_het))`) and Ts/Tv is
normal (`r round(sv("I20937", "ts_tv"), 2)`). Capture is typical for its batch
(705: `r round(sv("I20937", "pct_on_target"), 1)`% on target, rank
`r rank(b705$pct_on_target)[b705$sample == "I20937"]` of `r nrow(b705)`), but its
error rate (`r signif(sv("I20937", "error_rate"), 3)`) and missingness
(`r sprintf("%.1f%%", sv("I20937", "mean_pct_N_final"))`) are the highest in the
panel, consistent with reads diverging from the reference. 

## Samples to watch

```{r}
ab_sample <- ab_final |>
  summarize(
    n_evaluated = sum(status != "exempt"),
    pct_blanked = 100 * sum(status == "blanked") / n_evaluated,
    median_ab = median(ab, na.rm = TRUE),
    .by = sample
  )

flags <- sample_final |>
  left_join(select(ab_sample, sample, pct_blanked, median_ab), by = "sample") |>
  mutate(
    `low depth` = mean_dp < quantile(mean_dp, 0.05),
    `high missingness` = mean_pct_N_final > quantile(mean_pct_N_final, 0.95),
    `low Ts/Tv` = ts_tv < quantile(ts_tv, 0.05),
    `many singletons` = n_singletons > quantile(n_singletons, 0.95),
    `high duplication` = pct_duplication > quantile(pct_duplication, 0.95),
    `low mapping` = pct_mapped < quantile(pct_mapped, 0.05),
    # From the allele-balance section: many hets outside the 02b window, and a
    # balance skewed toward the reference, the signature of a second DNA source
    `many hets blanked` = pct_blanked > quantile(pct_blanked, 0.95),
    `low allele balance` = median_ab < quantile(median_ab, 0.05),
    .by = seq_batch
  ) |>
  # Not batch-relative: the confidence flag comes from 04b, which already
  # z-scores within the assigned sex group
  mutate(`sex not confident` = !sex_confident) |>
  mutate(n_flags = rowSums(pick(`low depth`:`sex not confident`)))

watch <- flags |>
  filter(n_flags >= 2) |>
  arrange(desc(n_flags))

# Low allele balance against sex-call confidence, for the text below
ab_sex <- with(flags, table(
  low_ab = factor(`low allele balance`, levels = c(FALSE, TRUE)),
  not_confident = factor(`sex not confident`, levels = c(FALSE, TRUE))
))
ab_sex_p <- fisher.test(ab_sex)$p.value

watch |>
  select(sample, species, seq_batch, n_flags, mean_dp, mean_pct_N_final, ts_tv,
         n_singletons, pct_duplication, pct_mapped, pct_blanked, median_ab) |>
  qc_gt()
```

Which checks each listed sample failed:

```{r}
watch |>
  select(sample, n_flags, `low depth`:`sex not confident`) |>
  pivot_longer(!c(sample, n_flags), names_to = "check") |>
  filter(value) |>
  summarize(failed = paste(check, collapse = ", "), .by = c(sample, n_flags)) |>
  arrange(desc(n_flags)) |>
  qc_gt(full_width = TRUE)
```

### Four additional samples were flagged contaminant

Four samples not flagged as contaminated in the metadata failed these checks in a
way consistent with a second source of DNA, and were removed in 03c
(`metadata/contam_samples_qc.txt`). All four had low allele balance:

- **I1495 and I1507** are females whose Z depth ratio is well above the female
  norm (`r round(sv("I1495", "z_depth_ratio"), 2)` and
  `r round(sv("I1507", "z_depth_ratio"), 2)`, against a female median of
  `r round(median(sample_final$z_depth_ratio[sample_final$sex == "female"]), 2)`),
  which is what reads from a male would add.
- **I1624** is a male flagged on both its depth ratio and its heterozygous
  gametolog sites.
- **I1517** is a male with a confident sex call, but a median allele balance of
  `r round(sv("I1517", "ab_median"), 2)`.
  **I1671** fit same pattern more weakly and was kept: a female with a raised depth ratio
  (`r round(sv("I1671", "z_depth_ratio"), 2)`) and low balance for its batch.

## All flagged contaminant samples

These are the samples flagged as contaminated in the metadata (Felipe's assessment)
plus the four mentioned above.

```{r}
# Heterozygosity is 05a's pooled `het_pct_final` here, since `all_final` only
# holds the retained samples
flagged_cmp <- per_sample |>
  filter(retained | contam %in% TRUE) |>
  mutate(group = ifelse(retained, "retained", "flagged")) |>
  select(!c(n_singletons, sex_confident)) |>
  left_join(select(vcf_stats_flagged, sample, n_singletons), by = "sample") |>
  left_join(transmute(sex_flagged, sample, sex_confident = confident == "yes"),
            by = "sample")
flagged_cmp |>
  summarize(
    n = n(),
    across(c(mean_dp, mean_pct_N_final, het_pct_final, ab_median,
             ab_pct_below_0.35, ab_pct_blanked, n_singletons, ts_tv),
           \(x) median(x, na.rm = TRUE)),
    pct_sex_not_confident = 100 * mean(!sex_confident, na.rm = TRUE),
    .by = group
  ) |>
  qc_gt(title = "Medians, flagged vs. retained samples", full_width = TRUE)
```

<hr style="height:1pt; visibility:hidden;" />

```{r}
# Interactive, with the sample ID as the tooltip
p <- flagged_cmp |>
  ggplot(aes(x = ab_median, y = het_pct_final, color = group)) +
  geom_point_interactive(aes(tooltip = sample, data_id = sample),
                         size = 2.5, alpha = 0.8) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Allele balance vs. heterozygosity",
    x = "Median allele balance of evaluated hets",
    y = "Percentage of called sites that are heterozygous",
    color = NULL
  )
girafe(ggobj = p, width_svg = 8, height_svg = 5)
```

`r n_low_ab_flagged` of the `r n_contam` flagged samples have a median allele
balance below `r ab_low`, against `r n_low_ab_retained` of the
`r nrow(sample_final)` retained samples. The flagged samples with low balance or
a sex call that is not confident are below. 
Across all flagged samples,
`r sum(flagged_cmp$group == "flagged" & flagged_cmp$mean_pct_N_final > 10)` of
`r n_contam` have more than 10% missing data over the final loci, so the
missingness filter would likely have removed them anyway.

```{r}
flagged_cmp |>
  filter(group == "flagged", ab_median < ab_low | !sex_confident) |>
  select(sample, species, seq_batch, mean_dp, mean_pct_N_final, het_pct_final,
         ab_median, ab_pct_below_0.35, n_singletons, sex_confident) |>
  arrange(ab_median) |>
  qc_gt(digits = 3)
```

## Heterozygosity per sample

```{r}
het_per_sample |>
  ggplot(aes(x = mean_dp, y = pct_het, color = species)) +
  geom_point(size = 2.5, alpha = 0.8) +
  labs(
    title = "Depth vs. heterozygosity",
    x = "Mean depth of coverage",
    y = "Percentage of called sites that are heterozygous",
    color = "Species"
  ) +
  theme(legend.text = element_text(face = "italic"))
```

## Genotype counts by group

```{r}
geno_labels <- c(
  n_het = "Heterozygous",
  n_alt = "Hom. non-ref.",
  n_diff = "Diff. from ref."
)

geno_r <- geno_smp |>
  select(sample, genotype, geno_pct) |>
  pivot_wider(names_from = genotype, values_from = geno_pct) |>
  with(round(cor(n_het, n_alt), 2))

# One boxplot per group, with the three metrics stacked as rows. Several groups
# have fewer than five samples, so the group size goes into the axis label, and
# it is counted per facet panel so that a group split across panels is not
# labeled with its total
# Labels and title are markdown (ggtext), so that species names can be italic
geno_plot <- function(df, x, xlab, title, facet = NULL, italic_x = FALSE) {
  df |>
    mutate(
      genotype = factor(genotype, levels = names(geno_labels), labels = geno_labels),
      group = paste0(
        if (italic_x) paste0("*", .data[[x]], "*") else .data[[x]],
        "<br>(", n_distinct(sample), ")"
      ),
      .by = all_of(c(x, facet))
    ) |>
    # Keep the order of `x` rather than sorting the pasted labels
    mutate(group = fct_reorder(group, as.integer(factor(.data[[x]])))) |>
    ggplot(aes(x = group, y = geno_pct)) +
    geom_boxplot(outlier.shape = NA, fill = "grey90", width = 0.6) +
    geom_jitter(aes(color = seq_batch), width = 0.15, height = 0, size = 2, alpha = 0.7) +
    facet_grid(
      rows = vars(genotype),
      cols = if (!is.null(facet)) vars(.data[[facet]]),
      scales = "free",
      space = "free_x"
    ) +
    scale_color_brewer(palette = "Set2") +
    labs(
      title = title,
      x = xlab,
      y = "Percentage of called sites",
      color = "Batch"
    ) +
    # theme_bw() draws a border around each facet panel
    theme_bw(base_size = 14) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      legend.position = "top",
      axis.text.x = element_markdown(),
      plot.title = element_markdown(face = "bold", hjust = 0.5)
    )
}
```

Three per-sample metrics (each a percentage of called sites averaged over the
final loci):

- **Heterozygous**: heterozygous sites.
- **Hom. non-ref.**: sites homozygous for a non-reference allele.
- **Diff. from ref.**: homozygous non-reference sites plus half the heterozygous
  sites, i.e. the share of sites at which a single allele copy differs from the
  reference.

### By species and land type

*B. insularis* and *B. sazimai* were sequenced only in the 605 batch and most
*B. alcatraz* in the 705 batch (see *Sequencing batch*). *B. germanoi* and
*B. jabrensis* have one sample each, and the *jabrensis* sample, I20937, is
expected to look diverged and homozygous (see *The singleton outlier: I20937*).

```{r}
#| fig-height: 8
geno_plot(geno_smp, x = "species", facet = "land_type", italic_x = TRUE,
          xlab = "Species", title = "Genotype counts by species and land type")
```

### *B. jararaca* by state

States are ordered from north to south. Batch is confounded with state (see
*Heterozygosity and sequencing batch*), so a difference between states that
matches the batch colors cannot be attributed to either.

```{r}
#| fig-height: 8
geno_smp |>
  filter(species == "jararaca") |>
  mutate(state = factor(state, levels = c("BA", "MG", "ES", "RJ", "SP", "PR", "SC", "RS"))) |>
  geno_plot(x = "state", xlab = "State", title = "*B. jararaca* genotype counts by state")
```

<details><summary>R session info</summary>

```{r}
#| echo: true
sessionInfo()
```

</details>
