---
title: "Locus and sample QC"
author: "Jelmer Poelstra"
date: today
format:
  html:
    self-contained: true
    cache: false
    number-sections: true
    number-depth: 2
    toc: true
    toc-depth: 3
    fig-dpi: 300
    df-print: kable
    grid:
      body-width: 800px
knitr:
  opts_chunk:
    message: FALSE
    out.width: "85%"
    R.options:
      knitr.table.html.attr: "quarto-disable-processing=true"
execute:
  echo: false
  warning: false
editor_options:
  chunk_output_type: console
---

-------

<br>

```{r}
# --- Setup
# Load packages
library(tidyverse)
library(gt)
library(here)

# Settings
theme_set(theme_classic(base_size = 14))
theme_update(plot.title = element_text(face = "bold", hjust = 0.5))

# ggplot picks a plot's default scale by looking up `scale_<aes>_<type>` in the
# global environment before its own namespace, so shadowing it here doubles the
# x-axis break density everywhere at once. Plots that pass their own `breaks`
# still win, since `breaks` takes precedence over `n.breaks`.
scale_x_continuous <- function(...) ggplot2::scale_x_continuous(n.breaks = 10, ...)

# --- Table helpers
# Columns whose values are all whole numbers are printed without decimals;
# everything else numeric gets `digits` decimals, so no table shows 5+ decimals
is_whole <- function(x) is.numeric(x) && all(x == round(x), na.rm = TRUE)

# Tables are left at their natural width unless `full_width` is set, since most
# of them have a handful of narrow columns and look better not stretched out
qc_gt <- function(df, digits = 2, title = NULL, full_width = FALSE) {
  tab <- df |>
    gt() |>
    fmt_number(
      columns = where(\(x) is.numeric(x) && !is_whole(x)),
      decimals = digits
    ) |>
    fmt_integer(columns = where(is_whole)) |>
    sub_missing(missing_text = "–") |>
    cols_label_with(fn = \(x) str_replace_all(x, "_", " ")) |>
    tab_options(
      table.width = if (full_width) pct(100) else "auto",
      table.font.size = px(15),
      data_row.padding = px(3),
      column_labels.font.weight = "bold",
      heading.title.font.weight = "bold"
    )
  if (!is.null(title)) tab <- tab |> tab_header(title = title)
  tab
}

# For the descriptive pipeline tables, whose cells contain inline code
doc_gt <- function(df, full_width = TRUE) {
  df |>
    gt() |>
    fmt_markdown(columns = where(is.character)) |>
    sub_missing(missing_text = "–") |>
    cols_label_with(fn = \(x) str_replace_all(x, "_", " ")) |>
    tab_options(
      table.width = if (full_width) pct(100) else "auto",
      table.font.size = px(15),
      data_row.padding = px(3),
      column_labels.font.weight = "bold"
    )
}

# Define input files (all written by scripts/05a_prep-qc-tables.R)
indir <- here("results/stats/summaries")
all_file <- file.path(indir, "sample-by-locus-stats.tsv")
locus_file <- file.path(indir, "per-locus-stats.tsv")
sample_file <- file.path(indir, "per-sample-stats.tsv")
summary_file <- file.path(indir, "filtering-summary.tsv")

# Written by 04g, not 05a: depth against distance from the capture-probe edge
profile_file <- here("results/stats/depth_profile/depth_profile.tsv")

# Written by the 4I side runs of 04b and 04f, which scored the flagged samples
# together with the retained ones
sex_flagged_file <- here("results/stats/sex_z_flagged/sex_calls.tsv")
vcf_stats_flagged_file <- here("results/stats/vcf_tstv_flagged/per_sample_vcf_stats.tsv")

# Written by 04h: every het genotype that reached the 02b allele-balance filter
ab_files <- list.files(here("results/stats/allele_balance"),
                       pattern = "_het-ab\\.tsv\\.gz$", full.names = TRUE)

# Read input files. `z_class` is NA for every autosomal locus, which is the
# large majority, so readr guesses logical from the leading rows and silently
# turns each real class into NA unless the type is given.
all <- read_tsv(all_file, col_types = cols(z_class = "c", .default = col_guess()))
per_locus <- read_tsv(locus_file, show_col_types = FALSE)
per_sample <- read_tsv(sample_file, show_col_types = FALSE)
filtering_summary <- read_tsv(summary_file, show_col_types = FALSE)
dp_profile <- read_tsv(profile_file, show_col_types = FALSE)
sex_flagged <- read_tsv(sex_flagged_file, show_col_types = FALSE)
vcf_stats_flagged <- read_tsv(vcf_stats_flagged_file, show_col_types = FALSE)
# `ao` is '.' where freebayes reports no alternate observation
het_ab <- read_tsv(ab_files, id = "sample", na = c("NA", "."),
                   col_types = "cicdddnc") |>
  mutate(sample = str_remove(basename(sample), "_het-ab\\.tsv\\.gz$"))

# The sample-by-locus table still holds the loci dropped by 04b/04c, so that the
# sections below can compare them against the retained ones; everything that is
# meant to describe the final data set filters on `retained_final` first
all_final <- all |> filter(retained_final)
locus_final <- per_locus |> filter(retained_final)
sample_final <- per_sample |> filter(retained)

n_contam <- sum(per_sample$contam, na.rm = TRUE)

# Samples with a median allele balance below this are listed in the
# flagged-samples section and counted in the summary
ab_low <- 0.45
n_low_ab_flagged <- sum(per_sample$contam %in% TRUE & per_sample$ab_median < ab_low, na.rm = TRUE)
n_low_ab_retained <- sum(sample_final$ab_median < ab_low, na.rm = TRUE)
```

## Summary

- `r nrow(sample_final)` of `r nrow(per_sample)` samples and
  `r nrow(locus_final)` of `r nrow(per_locus)` loci are retained.
- Sequencing batch is the largest technical factor: capture worked about twice as
  well in the 605 run, so depth-based numbers should be read by batch.
- Short (~200 bp) baits have about half the depth of long ones because of capture
  geometry, but give only modestly worse usable data.
- The 02b allele-balance filter blanks ~2% of evaluated hets; the pooled balance
  shows reference bias, not a genome-wide paralog signal.
- Paralog filters: `r sum(per_locus$stage == "dropped: excess het")` loci are
  dropped for excess heterozygosity and
  `r sum(per_locus$stage == "dropped: allele balance")` on allele balance alone;
  all `r sum(str_starts(per_locus$stage, "dropped: Z"))` sex-linked loci are
  dropped.
- I1495, I1507, I1624 and I1517 look like they may contain a second DNA source;
  none is flagged in the metadata.
- Of the `r n_contam` metadata-flagged samples, `r n_low_ab_flagged` have a median
  allele balance below `r ab_low`, against `r n_low_ab_retained` of the retained
  samples.
- I20937, the only *B. jabrensis*, has 10% of all singletons.

## Pipeline overview

### Step 1 --- Locus definition by BLAST

Probes are BLASTed against the reference genome (`01b_blast.sh`) and the hits
are filtered (`01d_filter-loci.R`):

```{r}
tribble(
  ~Step, ~Setting, ~Value,
  "BLAST", "`--top_n_query`", "5 hits per probe",
  "BLAST", "`--pct_id`", "≥ 90%",
  "BLAST", "`--pct_qcov` (HSP-based)", "≥ 80%",
  "`01d_filter-loci.R`", "Percent identity (`pident`)", "≥ 95% for at least one hit",
  "`01d_filter-loci.R`", "Query coverage (`qcovhsp`)", "≥ 90% for at least one hit",
  "`01d_filter-loci.R`", "Multi-hit probes", "Removed entirely (probes with > 1 BLAST hit)",
  "BED output", "Buffer around each hit", "300 bp on each side (`buffer_size`)"
) |>
  doc_gt()
```

Result: 2,000 probes → 1,936 with BLAST hits → 1,819 loci.

### Step 2 --- Genotype calling and consensus generation

A) Read processing and variant calling ---
   uses the nf-core Sarek pipeline, run per sample:

```{r}
tribble(
  ~Step, ~Tool, ~Setting,
  "Trimming", "fastp 0.24.0", "Adapter trimming, remove reads <15 bp, otherwise fastp defaults",
  "Mapping", "bwa-mem 0.7.18", "Default settings",
  "Duplicates", "GATK4 MarkDuplicates", "Duplicates marked — these are then ignored by freebayes",
  "Base recalibration", "—", "Skipped (`skip_tools: baserecalibrator`)",
  "Calling regions", "—", "Restricted to the probe BED (`intervals`)",
  "Variant calling", "freebayes 1.3.10", "`--report-monomorphic --genotype-qualities`, otherwise default"
) |>
  doc_gt()
```

B) Genotype filtering and consensus generation ---
   eight `bcftools` steps in `02b_bcftools-consensus.sh`:

```{r}
tribble(
  ~Step, ~Setting, ~Value, ~Effect,
  "0\\. Block splitting", "—", "—", "Multi-bp monomorphic records split into one per base",
  "1\\. Genotype quality", "`--filter_expr`", "`FORMAT/DP < 5 || FORMAT/GQ < 20`", "Genotypes below depth/GQ set to missing (`./.`)",
  "2\\. Site quality", "`--min_qual`", "30", "Genotypes at low-QUAL variant sites set to missing",
  "3\\. Indel proximity", "`--snp_gap`", "5 bp", "Genotypes at SNPs within 5 bp of an indel set to missing",
  "4\\. Normalization", "`bcftools norm`", "`--multiallelics -any --atomize`", "Splits multiallelics and MNVs, left-aligns indels",
  "5\\. Indels", "`bcftools view`", "`-V indels`", "Indels removed; SNPs only",
  "6\\. Allele balance", "`--min_ab` / `--max_ab`", "0.25 / 0.75", "Skewed heterozygote genotypes set to missing (only where AO + RO is within 2 reads of DP, `--ab_tol`)",
  "7\\. Multiallelics", "`--drop_multiallelic`", "`true`", "Positions left with >1 record after step 4 are removed"
) |>
  doc_gt()
```

_(Step 0 exists because freebayes occasionally merges a run of invariant positions_
_into a single record (REF several bp long, ALT `.`), which can span past the end_
_of the per-region reference slice given to `bcftools consensus` and crash it.)_

Consensus sequences are then generated with `bcftools consensus`, where
heterozygous sites get IUPAC ambiguity codes and missing/absent genotypes
become `N`.

### Step 3 --- Pivot, trim, and filter consensus sequences

A) Pivot to locus-centric FASTA files (`03_pivot-fasta.sh`).

B) Per-locus trimming with ClipKIT --- `03b_clipkit.sh` to remove the terminal
   `N` stretches caused by the buffers (terminal column is trimmed if >25% of
   sequences have `N` at that position).

C) Sample and locus filtering --- `03c_filter-loci.sh`, applied in this order:

```{r}
tribble(
  ~Step, ~Parameter, ~Value, ~Effect,
  "1\\. A priori sample removal", "`--remove_samples`", "—", "Samples flagged as contaminated in `metadata/metadata-final.tsv`",
  "2\\. Per-sample missingness", "`max_missing_sample`", "0.1", "Samples with mean >10% `N` across all loci removed",
  "3\\. Minimum locus length", "`min_locus_length`", "150 bp", "Post-trimming alignment length",
  "4\\. Per-locus missingness", "`max_missing_locus`", "0.2", "Loci with mean >20% `N` across retained samples removed"
) |>
  doc_gt()
```

_(Per-locus missingness in step 4 is recomputed after sample removal in steps_
_1-2, so the locus filter is based only on retained samples. VCFs are then subset_
_to the retained samples and loci.)_

### Step 4 --- Sex-linked loci

Sample sex is called from the Z:autosome depth ratio, and each Z locus is
classified as hemizygous, PAR-like, gametolog, or paralog, using the same
per-site excess-heterozygosity statistic as step 5. **Every** sex-linked locus
is currently dropped (`z_policy=all`).

### Step 5 --- Paralog filter: excess heterozygosity and allele balance

Paralogous loci are detected using two tests; a locus failing either is dropped:

- The first test uses a per-site statistic `f`, the fraction of called samples that
  are heterozygous.
- The second uses allele balance: where one of two collapsed copies carries a variant, a
  het shows it on a quarter to a third of the reads rather than half, in most
  samples that are heterozygous there.

```{r}
tribble(
  ~Parameter, ~Value, ~Effect,
  "`min_call_frac`", "0.25", "A site is only scored when at least this fraction of samples has a called genotype",
  "`max_het_frac`", "0.6", "A scored site with `f` above this is an 'excess-het site' (the HWE maximum is 0.5)",
  "`max_site_frac`", "0.005", "A locus is dropped when more than this fraction of its scored sites are excess-het sites...",
  "`min_het_sites`", "3", "...and it has at least this many excess-het sites",
  "`min_ab_hets`", "5", "A sample counts toward a locus's allele-balance test when it has at least this many evaluated hets there...",
  "`ab_skew_cutoff`", "0.4", "...and the locus is skewed in that sample when their median allele balance is below this",
  "`min_ab_samples`", "20", "A locus is only judged on allele balance with at least this many counted samples",
  "`max_ab_skew_frac`", "0.5", "A locus is dropped when it is skewed in more than this fraction of counted samples"
) |>
  doc_gt()
```

## Filtering summary

Loci and samples remaining after each filtering step:

```{r}
filtering_summary |> qc_gt(full_width = TRUE)
```

Step at which each locus and sample was dropped:

```{r}
per_locus |> count(stage, name = "n_loci") |> qc_gt()
per_sample |> count(stage, name = "n_samples") |> qc_gt()
```

## Depth of coverage

### Per sample

The depth figures in this section are mean FORMAT/DP over called genotypes
within the retained locus intervals. These are a capture probe plus a 300-bp
buffer on each side, after ClipKIT trimmed the poorly covered ends. A median of
~350 bp of buffer remains on short loci and ~310 bp on long ones. The buffer is
much less deeply covered than the probe (~25x over its full width, against ~45x
on short and ~70x on long probes; see *Locus lengths* below). These numbers are
therefore lower than on-probe depth, and loci of different lengths are not
directly comparable. Between samples the effect is small: buffer depth is
0.36--0.48 of probe depth in every sample.

```{r}
sample_final |>
  ggplot(aes(x = mean_dp)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black") +
  labs(
    title = "Mean depth of coverage per sample",
    x = "Mean depth of coverage",
    y = "Frequency (samples)"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))
```

#### The depth outlier: I1767

I1767 has by far the highest depth. Looks like it was simply sequenced deeper,
as other stats for this sample are not unusual.
Duplication at 10.8% is the second-highest, but that makes sense given the depth.

```{r}
sample_final |>
  select(sample, species, mean_dp, n_reads, pct_on_target, enrichment,
         pct_duplication, error_rate, mean_pct_N_final, ts_tv, n_singletons) |>
  arrange(desc(mean_dp)) |>
  head(5) |>
  qc_gt(digits = 4)
```

### Per locus

```{r}
locus_final |>
  ggplot(aes(x = mean_dp)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "black") +
  labs(
    title = "Mean depth of coverage per locus",
    x = "Mean depth of coverage (averaged over samples)",
    y = "Frequency (loci)"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

# Depth is the main driver of missingness, so these should be related
locus_final |>
  ggplot(aes(x = mean_dp, y = 100 * (1 - mean_call_frac))) +
  geom_point(alpha = 0.3, size = 1.5, color = "steelblue") +
  labs(
    title = "Locus depth vs. locus missingness",
    x = "Mean depth of coverage",
    y = "Mean percentage of missing data (N)"
  )
```

## Mapping and capture

Per-sample stats:

```{r}
sample_final |>
  select(sample, n_reads, pct_mapped, pct_properly_paired, pct_mq0,
         pct_duplication, pct_on_target, target_mean_dp, enrichment) |>
  summarize(across(!sample, list(min = min, median = median, max = max))) |>
  pivot_longer(
    everything(),
    names_to = c("metric", "stat"),
    names_pattern = "(.*)_(min|median|max)"
  ) |>
  pivot_wider(names_from = stat, values_from = value) |>
  gt() |>
  # Read counts are in the millions and the rates are percentages, so a shared
  # format would render one of them unreadably
  fmt_number(columns = c(min, median, max), decimals = 2) |>
  fmt_integer(columns = c(min, median, max), rows = metric == "n_reads") |>
  cols_label_with(fn = \(x) str_replace_all(x, "_", " ")) |>
  tab_options(
    table.width = pct(100),
    table.font.size = px(15),
    data_row.padding = px(3),
    column_labels.font.weight = "bold"
  )
```

```{r}
# The share of all sequenced bases that landed inside the probe intervals. The
# targets are 2.2 Mb of a 1.57 Gb genome, so even a modest percentage here is a
# large enrichment. Filled by batch because the two runs separate almost
# completely on this axis -- see the section below.
sample_final |>
  ggplot(aes(x = pct_on_target, fill = seq_batch)) +
  geom_histogram(binwidth = 0.25, color = "black") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "On-target rate per sample",
    x = "Percentage of sequenced bases inside the probe intervals",
    y = "Frequency (samples)",
    fill = "Batch"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

# Read count sets the ceiling on depth, but the conversion rate differs per
# batch, so there are two trends rather than one. Coloring by duplicate rate
# here would be misleading: duplication is itself higher in the 605 batch, so
# the gradient would mostly re-encode the batch.
sample_final |>
  ggplot(aes(x = n_reads / 1e6, y = mean_dp, color = seq_batch)) +
  geom_point(size = 2.5, alpha = 0.8) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 0.7) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "Read count vs. depth",
    x = "Read count (millions)",
    y = "Mean depth of coverage",
    color = "Batch"
  )
```

### Sequencing batch

The samples were sequenced in two runs (script 02d reads the run and flowcell
from the FASTQ read names), and the capture worked about twice as well in one of
them. This is the single largest structuring factor in the coverage data, so it
is worth stating explicitly before any of the depth- or missingness-based
numbers below are read as biology.

```{r}
sample_final |>
  summarize(
    n_samples = n(),
    across(c(n_reads, pct_on_target, enrichment, pct_duplication, mean_dp,
             mean_pct_N_final),
           mean),
    dp_per_Mread = mean(mean_dp / (n_reads / 1e6)),
    .by = seq_batch
  ) |>
  arrange(seq_batch) |>
  relocate(dp_per_Mread, .after = mean_dp) |>
  qc_gt(title = "Per-batch means", full_width = TRUE)
```

The mechanism is the on-target rate, and nothing else: across all retained
samples, the depth a million reads buys correlates with `pct_on_target` at
r = 0.998, and still at 0.988 *within* each batch. The two batches do not
overlap at all on that axis, which is why the scatterplot above splits into two
lines instead of one cloud.

```{r}
sample_final |>
  summarize(
    n = n(),
    on_target_min = min(pct_on_target),
    on_target_max = max(pct_on_target),
    dup_min = min(pct_duplication),
    dup_max = max(pct_duplication),
    r_dp_vs_on_target = cor(mean_dp / (n_reads / 1e6), pct_on_target),
    r_dp_vs_dup = cor(mean_dp / (n_reads / 1e6), pct_duplication),
    .by = seq_batch
  ) |>
  arrange(seq_batch) |>
  qc_gt(title = "Spread and correlations within each batch", full_width = TRUE)

# Loci called per sample, per batch, for the text below
loci_called <- sample_final |>
  summarize(mean = round(mean(n_loci_called)), min = min(n_loci_called),
            .by = seq_batch)
lc <- function(b, col) loci_called[[col]][loci_called$seq_batch == b]
```

Two things partly offset the difference. The weaker-capture batch was sequenced
about 30% deeper, and it also has the lower duplicate rate, so its samples end
up down and to the right in the scatterplot rather than simply below. The net
effect on the final data is real but modest: mean missingness after filtering is
`r round(mean(sample_final$mean_pct_N_final[sample_final$seq_batch == "605_HCTGLBCXX"]), 2)`%
in the 605 batch against
`r round(mean(sample_final$mean_pct_N_final[sample_final$seq_batch == "705_H2HLLBCXY"]), 2)`%
in the 705 batch, and the number of loci called per sample is essentially
identical between them (mean `r lc("605_HCTGLBCXX", "mean")` and
`r lc("705_H2HLLBCXY", "mean")` of `r nrow(locus_final)`, with a per-sample
minimum of `r lc("605_HCTGLBCXX", "min")` and `r lc("705_H2HLLBCXY", "min")`).

Batch is not confounded with the main biological groupings: the split is close
to even for *jararaca* and for the mainland samples. It is not even, however,
for the island endemics --- *insularis* and *sazimai* were sequenced only in the
605 batch and most *alcatraz* samples in the 705 batch --- so any
between-species comparison of a coverage-derived quantity should be checked
against this table.

```{r}
sample_final |>
  count(land_type, species, seq_batch) |>
  pivot_wider(names_from = seq_batch, values_from = n, values_fill = 0) |>
  arrange(land_type, species) |>
  qc_gt(title = "Samples per land type, species and batch")
```

One sample crosses over: `I1619` was sequenced in the 605 batch but sits with
the 705 cloud. It has the highest duplicate rate in the data set (14.4%) on top
of the lowest on-target rate of its own batch.

Batch also affects which samples survive filtering: of the
`r sum(per_sample$stage == "dropped: missingness")` samples dropped for
missingness in 03c,
`r sum(per_sample$stage == "dropped: missingness" & per_sample$seq_batch == "705_H2HLLBCXY", na.rm = TRUE)`
are from the 705 batch.

```{r}
per_sample |>
  filter(stage %in% c("retained", "dropped: missingness")) |>
  count(seq_batch, stage) |>
  pivot_wider(names_from = stage, values_from = n, values_fill = 0) |>
  qc_gt(title = "Samples retained and dropped for missingness, per batch")
```

## Locus lengths

```{r}
# All loci with a length from 03c, so the `min_locus_length` cutoff is visible
per_locus |>
  mutate(set = ifelse(retained_final, "retained", "dropped")) |>
  ggplot(aes(x = length, fill = set)) +
  geom_histogram(binwidth = 50, color = "black") +
  geom_vline(xintercept = 150, linetype = "dashed") +
  labs(x = "Locus length (bp)", y = "Frequency (loci)", fill = NULL) +
  scale_fill_manual(values = c(retained = "steelblue", dropped = "grey70")) +
  # 125-bp breaks collide on an axis this wide, so this plot keeps 250
  scale_x_continuous(breaks = seq(0, 5000, by = 250)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))
```

Length and depth go together
(r = `r round(cor(locus_final$length, locus_final$mean_dp, use = "complete.obs"), 2)`),
for two reasons.

The first is capture geometry. A locus is one bait plus 300 bp of buffer on each
side, and the baits come in two sizes, ~200 and ~1,950 bp. The buffer is covered
at ~25x and makes up three quarters of a short interval but under a quarter of a
long one. On the bait itself, capture efficiency ramps up over the first ~300 bp
from its edge (script 04g, below), so a ~200-bp bait never reaches the plateau
that long baits hold.

```{r}
# Probe length class per locus, also used in the allele-balance section
probe_dp <- locus_final |>
  mutate(
    probe_len = n_sites_pre - 600,
    probe_class = ifelse(probe_len < 600, "short", "long")
  )

# The outermost bins on each side are thin -- few probes are long enough to fill
# the innermost bin of their class, and few loci sit far enough from a neighbor
# to fill the outermost buffer bin -- and they run against the trend, so bins
# holding under 10% of their class's typical bin are dropped rather than plotted
dp_profile_plot <- dp_profile |>
  filter(n_pos >= 0.1 * median(n_pos), .by = probe_class)

dp_profile_plot |>
  filter(dist_bin >= -300, dist_bin <= 600) |>
  ggplot(aes(x = dist_bin + 25, y = mean_dp, color = probe_class)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  labs(
    title = "Depth against distance from the capture-probe edge",
    subtitle = "Negative distances are buffer; 0 is the first on-probe base",
    x = "Distance from nearest probe edge (bp)",
    y = "Mean depth of coverage",
    color = "Probe class"
  ) +
  scale_x_continuous(breaks = seq(-300, 600, by = 100)) +
  scale_y_continuous(n.breaks = 8)
```

The second is trimming, where depth sets length rather than the reverse: ClipKIT
removes mostly-`N` ends, so a poorly covered locus finishes short (`pct_trimmed`
vs. depth
r = `r round(cor(locus_final$pct_trimmed, locus_final$mean_dp, use = "complete.obs"), 2)`).
Within the short class, where every locus starts from the same 800-bp interval,
final length still tracks depth
(r = `r round(with(filter(probe_dp, probe_class == "short"), cor(length, mean_dp)), 2)`).

```{r}
per_locus |>
  filter(retained_final) |>
  ggplot(aes(x = length, y = mean_dp, color = pct_trimmed)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "Locus length vs. depth",
    x = "Locus length after trimming (bp)",
    y = "Mean depth of coverage",
    color = "% trimmed"
  ) +
  scale_y_continuous(n.breaks = 8) +
  scale_color_viridis_c(option = "plasma", end = 0.9)
```

```{r}
# Short vs. long final loci, for the text below
by_len <- locus_final |>
  mutate(size = ifelse(length < 1000, "short", "long")) |>
  summarize(
    n = n(),
    dp = median(mean_dp),
    call = median(mean_call_frac),
    worst_call = median(min_call_frac),
    prop_var = median(prop_variable),
    n_pi = median(n_pi),
    .by = size
  )
bl <- function(size, col) by_len[[col]][by_len$size == size]
```

Median depth is `r round(bl("short", "dp"), 1)` for loci under 1,000 bp
(n = `r bl("short", "n")`) against `r round(bl("long", "dp"), 1)` for the longer
ones (n = `r bl("long", "n")`), but that gap overstates the difference in usable
data: the low-depth buffer is largely what ClipKIT trims away. After trimming,
the short loci are only modestly worse --- median call fraction
`r round(bl("short", "call"), 3)` against `r round(bl("long", "call"), 3)`, and
`r round(bl("short", "worst_call"), 2)` against
`r round(bl("long", "worst_call"), 2)` for the worst sample at each locus --- and
their per-base informativeness is similar (median `prop_variable`
`r round(bl("short", "prop_var"), 3)` against `r round(bl("long", "prop_var"), 3)`).
They contribute fewer parsimony-informative sites (median `r round(bl("short", "n_pi"))`
against `r round(bl("long", "n_pi"))`) because they are shorter, not because they are
called less well.

## End trimming

ClipKIT (script 03b) removed terminal columns that were mostly N, and script 03d
tabulated how much went per locus. The buffer is at most 75% of an interval (600
of 800 bp for a short locus), so a locus trimmed by 80-99% also lost part of its
probe: what remains is short and usually poorly covered.

Loci where ClipKIT failed (`trim_failed`) were left untrimmed and show up at 0%;
it failed because missing data was high across the whole locus.

```{r}
per_locus |>
  mutate(set = case_when(
    trim_failed ~ "ClipKIT failed",
    retained_final ~ "retained",
    .default = "dropped"
  )) |>
  ggplot(aes(x = pct_trimmed, fill = set)) +
  geom_histogram(binwidth = 2, color = "black") +
  labs(
    title = "Percentage of each alignment trimmed",
    x = "Percentage of alignment columns trimmed",
    y = "Frequency (loci)",
    fill = NULL
  ) +
  scale_fill_manual(values = c(retained = "steelblue", dropped = "grey70",
                               `ClipKIT failed` = "firebrick")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

# Heavy trimming should track with low retained depth, since both follow from a
# locus that captured poorly
locus_final |>
  ggplot(aes(x = pct_trimmed, y = mean_dp)) +
  geom_point(alpha = 0.3, size = 1.5, color = "steelblue") +
  labs(
    title = "Trimming vs. depth",
    x = "Percentage of alignment columns trimmed",
    y = "Mean depth of coverage"
  )
```

## Missingness

### Per locus

Missingness is shown here for every locus that entered step 3, including the
ones later dropped, so that the `max_missing_locus` threshold (20%) is visible
instead of cutting the distribution off at its own value. Loci are colored by
where they left the pipeline, since the ones dropped for length, sex linkage, excess
heterozygosity or allele balance are not a comment on missingness:

```{r}
#| fig-height: 4
per_locus |>
  ggplot(aes(x = pct_N_prefilt, fill = stage)) +
  geom_histogram(binwidth = 1, color = "black", linewidth = 0.2) +
  geom_vline(xintercept = 20, linetype = "dashed", color = "grey30") +
  labs(
    title = "Mean missingness per locus, before filtering",
    x = "Mean percentage of missing data (N)",
    y = "Frequency (loci)",
    fill = NULL
  ) +
  scale_fill_viridis_d(option = "turbo") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))
```

### Per sample

```{r}
# `pct_N` is 03c's value over all loci before the locus filters, so the samples
# it removed show up against the `max_missing_sample` cutoff
per_sample |>
  filter(!is.na(pct_N)) |>
  mutate(set = ifelse(retained, "retained", "dropped")) |>
  ggplot(aes(x = pct_N, fill = set)) +
  geom_histogram(binwidth = 1, color = "black", linewidth = 0.2) +
  geom_vline(xintercept = 10, linetype = "dashed", color = "grey30") +
  labs(
    title = "Mean missingness per sample, before filtering",
    x = "Mean percentage of missing data (N)",
    y = "Frequency (samples)",
    fill = NULL
  ) +
  scale_fill_manual(values = c(retained = "steelblue", dropped = "grey70")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

# Depth is the main driver here too
sample_final |>
  ggplot(aes(x = mean_dp, y = mean_pct_N_final)) +
  geom_point(alpha = 0.6, size = 2.5, color = "steelblue") +
  labs(
    title = "Sample depth vs. sample missingness",
    x = "Mean depth of coverage",
    y = "Mean percentage of missing data (N)"
  )
```

## Sample sex

Snakes are ZW, so females are hemizygous over the differentiated part of the Z
and show about half the male depth there.

```{r}
sample_final |>
  ggplot(aes(x = z_depth_ratio, fill = sex)) +
  geom_histogram(binwidth = 0.01, color = "black") +
  labs(
    title = "Z:autosome depth ratio",
    x = "Z:autosome depth ratio",
    y = "Frequency (samples)",
    fill = "Sex"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

sample_final |> count(sex, sex_confident) |> qc_gt()
```

Sex is assigned by splitting the sorted depth ratios at their largest internal
gap, so every sample gets a call however intermediate its ratio is. The
confidence flag is what marks the shaky ones: three statistics --- the depth
ratio, heterozygosity over the hemizygous Z, and the number of heterozygous
gametolog sites --- are z-scored within the assigned sex group using the median
and a MAD-based SD, and a sample exceeding `max_group_z` (3) on any of them is
flagged, as is any sample below `min_auto_dp` (15) autosomal depth.

Samples whose sex could not be called with confidence:

```{r}
sample_final |>
  filter(!sex_confident) |>
  select(sample, species, mean_dp, z_depth_ratio, z_het_density,
         n_het_gametolog, sex, sex_flags) |>
  arrange(z_depth_ratio) |>
  qc_gt(digits = 4)
```

Most flagged samples are females whose depth ratio sits between the two modes,
which the gap split has to assign to one side; the `sex_flags` column shows which
statistic flagged each sample.

## Excess heterozygosity

This is one of the two paralog tests in 04c (the other is allele balance,
below). A locus is dropped as a likely collapsed paralog when more than 0.5% of its
scored sites (`max_site_frac`), and at least 3 sites (`min_het_sites`), are
heterozygous in more than 60% of called samples (`max_het_frac`, against a
maximum of 0.5 under HWE). For those thresholds to be meaningful the flagged
loci should stand apart rather than form the tail of a continuous distribution.

Note that the `pct_max_site_het` column below is the *observed* maximum across
the sites of a locus, not the threshold.

```{r}
per_locus |>
  filter(!is.na(excess_site_frac)) |>
  mutate(set = ifelse(retained_final, "retained", "dropped")) |>
  ggplot(aes(x = 100 * excess_site_frac, fill = set)) +
  geom_histogram(binwidth = 0.05, color = "black") +
  # The 0.5% cutoff. It is not on its own a boundary, since dropping also needs
  # >= 3 excess-het sites, but no retained locus happens to sit right of it
  geom_vline(xintercept = 0.5, linetype = "dashed") +
  labs(
    title = "Excess-heterozygosity sites per locus",
    x = "Percentage of scored sites heterozygous in > 60% of samples",
    y = "Frequency (loci)",
    fill = NULL
  ) +
  scale_fill_manual(values = c(retained = "steelblue", dropped = "grey70")) +
  # Default breaks are spaced for a linear axis and leave the log1p scale with
  # almost none, so they are given explicitly
  scale_y_continuous(
    transform = "log1p",
    breaks = c(0, 1, 3, 10, 30, 100, 300, 1000),
    expand = expansion(mult = c(0, 0.1))
  )
```

The filter is not driven by one sequencing batch: all four dropped loci are
elevated in both, at 2-4x the batch's own panel mean (L1306 1.03/1.04, L1813
1.49/1.56, L663 1.40/1.38, L473 1.32/0.90 for 705/605, against panel means of
0.38 and 0.42).

The loci dropped for excess heterozygosity, with the ones just below the
threshold for comparison. `excess_site_frac` and `het_density` are small
fractions, so these are shown as percentages:

```{r}
per_locus |>
  filter(!is.na(n_excess_sites)) |>
  mutate(
    pct_excess_sites = 100 * excess_site_frac,
    pct_max_site_het = 100 * max_site_het_frac,
    pct_het_density = 100 * het_density
  ) |>
  select(locus, stage, length, n_sites_het_scored, n_excess_sites,
         pct_excess_sites, pct_max_site_het, pct_het_density) |>
  arrange(desc(n_excess_sites)) |>
  head(15) |>
  qc_gt()
```

## Allele balance

Step 6 of 02b sets a heterozygous genotype to missing when its allele balance,
AO / (AO + RO), falls outside 0.25-0.75. The filtered VCFs cannot show what that
removed, because a blanked genotype looks like any other missing one, so script
04h replays the 02b filter chain on the raw VCFs and records every het that
reaches step 6. It reproduces the filtered VCFs exactly (checked position by
position for one sample: every 'kept' and 'exempt' het is a het there, every
'blanked' one is missing).

Allele balance is a ratio, so unlike the depth figures above it does not depend
on where in the locus a site sits. Everything below is restricted to the retained
samples.

```{r}
ab_all <- het_ab |>
  filter(sample %in% sample_final$sample) |>
  left_join(select(per_locus, locus, stage, retained_final), by = "locus")

ab_final <- ab_all |> filter(retained_final)
ab_eval <- ab_final |> filter(status != "exempt")

ab_final |>
  count(status) |>
  mutate(pct = 100 * n / sum(n)) |>
  qc_gt(title = "Het genotypes at the allele-balance step, retained loci")

ab_exempt <- ab_final |>
  filter(status == "exempt") |>
  mutate(reason = case_when(
    is.na(ao) ~ "no AO reported",
    ro == 0 ~ "RO = 0 (split multiallelic)",
    .default = "AO + RO off DP by > 2 reads"
  )) |>
  count(reason) |>
  mutate(pct = 100 * n / sum(n)) |>
  arrange(desc(n))

ab_exempt |> qc_gt(title = "Why exempt hets were not evaluated")
```

'Exempt' hets are on records where AO + RO differs from DP by more than 2 reads
(`--ab_tol`), which 02b skips on purpose: after multiallelic splitting, one half
of a real 1/2 genotype shows no reference reads and would otherwise be thrown
out. That case is
`r sprintf("%.0f%%", ab_exempt$pct[ab_exempt$reason == "RO = 0 (split multiallelic)"])`
of exempt hets, and in total the filter does not look at
`r sprintf("%.1f%%", 100 * mean(ab_final$status == "exempt"))` of hets.

### Distribution

```{r}
ab_eval |>
  ggplot(aes(x = ab, fill = status)) +
  geom_histogram(binwidth = 0.02, boundary = 0, color = "black", linewidth = 0.2) +
  geom_vline(xintercept = c(0.25, 0.75), linetype = "dashed") +
  labs(
    title = "Allele balance of het genotypes before the filter",
    x = "Allele balance, AO / (AO + RO)",
    y = "Frequency (het genotypes)",
    fill = NULL
  ) +
  scale_fill_manual(values = c(kept = "steelblue", blanked = "grey70")) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.1)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

ab_eval |>
  mutate(depth = cut(dp, c(0, 10, 20, 40, 80, Inf),
                     labels = c("5-10", "11-20", "21-40", "41-80", "> 80"))) |>
  summarize(
    n_het = n(),
    median_ab = median(ab),
    pct_below_0.35 = 100 * mean(ab < 0.35),
    pct_above_0.65 = 100 * mean(ab > 0.65),
    pct_blanked = 100 * mean(status == "blanked"),
    .by = depth
  ) |>
  arrange(depth) |>
  qc_gt(title = "Evaluated hets by genotype depth, retained loci")
```

The distribution has a single mode at 0.5 and leans to the reference side:
`r sprintf("%.1f%%", 100 * mean(ab_eval$ab < 0.35))` of hets are below 0.35
against `r sprintf("%.1f%%", 100 * mean(ab_eval$ab > 0.65))` above 0.65. That
is the usual reference bias of mapping to a reference, which favors reads that
match it, and it is strongest at moderate depth. There is no second mode near
0.25, so paralogs are not a problem at the level of the whole data set; they show
up per locus (below). Overall, the filter removes
`r sprintf("%.1f%%", 100 * mean(ab_eval$status == "blanked"))` of evaluated hets.

### Per sample

```{r}
ab_sample <- ab_final |>
  summarize(
    n_het = n(),
    n_evaluated = sum(status != "exempt"),
    pct_blanked = 100 * sum(status == "blanked") / n_evaluated,
    pct_exempt = 100 * mean(status == "exempt"),
    median_ab = median(ab, na.rm = TRUE),
    pct_below_0.35 = 100 * mean(ab < 0.35, na.rm = TRUE),
    pct_above_0.65 = 100 * mean(ab > 0.65, na.rm = TRUE),
    .by = sample
  ) |>
  left_join(
    select(sample_final, sample, species, seq_batch, mean_dp, n_transitions, n_singletons),
    by = "sample"
  )

ab_sample |>
  ggplot(aes(x = pct_blanked, fill = seq_batch)) +
  geom_histogram(binwidth = 0.5, color = "black") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Share of evaluated hets blanked by the allele-balance filter",
    x = "Percentage of evaluated het genotypes blanked",
    y = "Frequency (samples)",
    fill = "Batch"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

ab_sample |>
  arrange(desc(pct_blanked)) |>
  head(8) |>
  select(sample, species, seq_batch, n_evaluated, pct_blanked, median_ab,
         pct_below_0.35, pct_above_0.65, mean_dp, n_transitions, n_singletons) |>
  qc_gt(title = "Samples with the most hets blanked")
```

The typical sample loses `r sprintf("%.1f%%", median(ab_sample$pct_blanked))` of
its evaluated hets (IQR
`r sprintf("%.1f-%.1f%%", quantile(ab_sample$pct_blanked, 0.25), quantile(ab_sample$pct_blanked, 0.75))`),
and the share is unrelated to depth (r =
`r round(cor(ab_sample$pct_blanked, ab_sample$mean_dp), 2)`). The samples well
above that come in two kinds:

- **Widened balance: I1768, I1670, I1684.** About a fifth of their hets are
  blanked, and their median balance is still close to 0.5. For I1768 and I1670
  both ends of the window lose hets about equally; I1684 loses more on the low
  side. Depth is not low (I1684 has the second-highest in the panel), and
  duplication, error rate, and insert size are ordinary. What the three share is
  a low variant count: they are among the samples with the
  fewest transitions and have very few singletons. A second DNA source would
  skew balance to one side, which the first two do not show, and the cause is
  not clear from the metrics in this report.
- **Skewed low: I1495, I1517, and less strongly I1624 and I1507.** The first two
  have a median balance below 0.4 with about 40% of their hets under 0.35; I1624
  and I1507 are at 0.43-0.44 with a quarter of their hets under 0.35. That is what
  a minor second source of DNA does: the contaminant's alleles appear as hets
  supported by a minority of reads. All four have more singletons than the median
  sample (`r median(sample_final$n_singletons)`), and three of them also fail sex
  calling (see *Samples to watch*). None is among the samples flagged as
  contaminated in the metadata.

### Per locus

Pooling all hets at a locus lets a few low-coverage samples dominate its share of
blanked hets, so loci are scored per sample instead. For each sample with at least
5 evaluated hets at a locus, the locus is 'skewed' in that sample if the median
balance is below 0.4. A collapsed paralog should be skewed in most samples that
are heterozygous there; a locus with a few noisy samples should not.

```{r}
min_ab_hets <- 5
min_ab_samples <- 20
ab_skew_cutoff <- 0.4
max_ab_skew_frac <- 0.5

ab_locus <- ab_all |>
  filter(status != "exempt") |>
  summarize(n_het = n(), median_ab = median(ab), .by = c(locus, stage, sample)) |>
  filter(n_het >= min_ab_hets) |>
  summarize(
    n_samples = n(),
    n_skewed = sum(median_ab < ab_skew_cutoff),
    pct_skewed = 100 * n_skewed / n_samples,
    .by = c(locus, stage)
  ) |>
  filter(n_samples >= min_ab_samples) |>
  left_join(
    ab_all |>
      filter(status != "exempt") |>
      summarize(n_het = n(), median_ab = median(ab),
                pct_blanked = 100 * mean(status == "blanked"), .by = locus),
    by = "locus"
  ) |>
  left_join(select(per_locus, locus, mean_dp, n_excess_sites), by = "locus")

# Pull one value for one locus, for the text
ab_l <- function(l, col) ab_locus[[col]][ab_locus$locus == l]

ab_locus |>
  summarize(
    n_loci = n(),
    median_pct_skewed = median(pct_skewed),
    max_pct_skewed = max(pct_skewed),
    .by = stage
  ) |>
  arrange(desc(median_pct_skewed)) |>
  qc_gt(title = paste0("Share of samples in which a locus is skewed, by locus fate (loci with >= ",
                       min_ab_samples, " counted samples)"))

ab_locus |>
  mutate(set = case_when(
    stage %in% c("retained", "dropped: allele balance") ~ stage,
    .default = "dropped: other"
  )) |>
  ggplot(aes(x = pct_skewed, fill = set)) +
  geom_histogram(binwidth = 2.5, boundary = 0, color = "black") +
  # The 04c cutoff (`max_ab_skew_frac`)
  geom_vline(xintercept = 100 * max_ab_skew_frac, linetype = "dashed") +
  labs(
    title = "Share of samples in which a locus has skewed allele balance",
    subtitle = paste0("Median balance < ", ab_skew_cutoff, " among samples with >= ",
                      min_ab_hets, " evaluated hets; loci with >= ",
                      min_ab_samples, " counted samples"),
    x = "Percentage of samples skewed",
    y = "Frequency (loci)",
    fill = NULL
  ) +
  scale_fill_manual(values = c(
    retained = "steelblue",
    `dropped: allele balance` = "firebrick",
    `dropped: other` = "grey70"
  )) +
  scale_y_continuous(
    transform = "log1p",
    breaks = c(0, 1, 3, 10, 30, 100, 300),
    expand = expansion(mult = c(0, 0.1))
  )

ab_locus |>
  arrange(desc(pct_skewed)) |>
  head(15) |>
  qc_gt(title = "Loci skewed in the largest share of samples")
```

Loci that 04b dropped as Z paralogs or gametologs, which this measure did not
select, also sit at the top, and the typical retained locus is skewed in
`r sprintf("%.0f%%", median(ab_locus$pct_skewed[ab_locus$stage == "retained"]))`
of samples, with 99% of retained loci at or below
`r sprintf("%.0f%%", quantile(ab_locus$pct_skewed[ab_locus$stage == "retained"], 0.99))`.

**L1826 is dropped by the allele-balance test alone.** It is skewed in
`r sprintf("%.0f%%", ab_l("L1826", "pct_skewed"))` of
`r ab_l("L1826", "n_samples")` samples, alongside L1839 (dropped as a Z paralog,
`r sprintf("%.0f%%", ab_l("L1839", "pct_skewed"))`) and L1813 (dropped for excess
heterozygosity, `r sprintf("%.0f%%", ab_l("L1813", "pct_skewed"))`). Like those
two, it has about twice the depth of a typical long locus
(`r round(ab_l("L1826", "mean_dp"), 0)`x, against a long-class median of
`r round(median(probe_dp$mean_dp[probe_dp$probe_class == "long"]), 0)`x), and its
`r ab_l("L1826", "n_excess_sites")` excess-het sites are too few for the
excess-het test. Balance near 0.25-0.35 at the same sites across dozens of
samples is what one variable copy of two looks like.

L10 is the most skewed retained locus
(`r sprintf("%.0f%%", ab_l("L10", "pct_skewed"))` of
`r ab_l("L10", "n_samples")` samples), but its depth is ordinary for a short
locus and it has no excess-het sites.

## Locus informativeness

The per-sample genotype counts in 05c measure how far each sample sits from the
reference, which says nothing about whether a locus separates the samples from
each other.
Script 04e counts, over the `r nrow(sample_final)` retained samples, the sites
that are variable among them and the subset that are parsimony-informative ---
where each of the two alleles is carried by at least two samples, so the site can
support a grouping.

```{r}
locus_final |>
  summarize(
    across(c(n_sites_scored, n_variable, n_pi, n_singleton), sum)
  ) |>
  mutate(
    pct_variable = 100 * n_variable / n_sites_scored,
    pct_pi = 100 * n_pi / n_sites_scored
  ) |>
  qc_gt()
```

```{r}
locus_final |>
  ggplot(aes(x = n_pi)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black") +
  labs(
    title = "Parsimony-informative sites per locus",
    x = "Number of parsimony-informative sites",
    y = "Frequency (loci)"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

locus_final |>
  ggplot(aes(x = 100 * prop_pi)) +
  geom_histogram(binwidth = 0.25, fill = "steelblue", color = "black") +
  labs(
    title = "Density of parsimony-informative sites per locus",
    x = "Percentage of scored sites that are parsimony-informative",
    y = "Frequency (loci)"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

# Longer loci carry more informative sites simply by being longer; loci far below
# the trend are conserved rather than badly sequenced
locus_final |>
  ggplot(aes(x = length, y = n_pi, color = mean_dp)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_viridis_c() +
  labs(
    title = "Locus length vs. informative sites",
    x = "Locus length (bp)",
    y = "Number of parsimony-informative sites",
    color = "Depth"
  )
```

## Heterozygosity per sample

Heterozygosity is the readout the downstream comparisons in 05c are built on, so
it is worth checking here that it is not being set by depth instead of by
biology. At low depth one allele of a true heterozygote can go unobserved, and
the genotype is then called homozygous --- allele dropout, which biases
heterozygosity down exactly where coverage is worst.

```{r}
het_per_sample <- all_final |>
  filter(genotype == "n_het") |>
  # Mean of per-locus percentages, as in 05c; 05a's `het_pct_final` pools sites
  # across loci instead, so it weights long loci more. NA where a locus is
  # uncalled in this sample
  summarize(pct_het = mean(geno_pct, na.rm = TRUE), .by = sample) |>
  left_join(sample_final, by = "sample")

het_per_sample |>
  ggplot(aes(x = pct_het, fill = sex)) +
  geom_histogram(binwidth = 0.02, color = "black", show.legend = FALSE) +
  # Stacking hides the shape of the smaller group, and the two are expected to
  # differ here, so each sex gets its own panel with a free y axis
  facet_wrap(vars(sex), ncol = 1, scales = "free_y") +
  labs(
    title = "Heterozygosity per sample",
    x = "Percentage of called sites that are heterozygous",
    y = "Frequency (samples)"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))
```

The two panels look the same, which is the point: with `z_policy=all` every
sex-linked locus is already gone, so female hemizygosity can no longer depress
female heterozygosity. Medians are 0.44% for females against 0.49% for males
across the whole panel, and 0.47% vs 0.49% within *B. jararaca* --- the residual
gap is the species composition, not the sex.

Both panels are bimodal in the same place, and the split follows species and
locality, not sex. The low cluster is every non-*jararaca* sample (medians
0.12-0.24%, all 20 of them below 0.40%) plus 29 *jararaca*; the high cluster is
*jararaca* only. Within *jararaca*, heterozygosity varies with sampling locality
(see *Heterozygosity and sequencing batch*).

```{r}
het_per_sample |>
  summarize(
    n = n(),
    median_pct_het = median(pct_het),
    min = min(pct_het),
    max = max(pct_het),
    .by = species
  ) |>
  arrange(median_pct_het) |>
  qc_gt(digits = 3)
```

Either slope is a warning, and they mean opposite things. Rising with depth
points to allele dropout, heterozygosity being lost where coverage is thin.
Falling with depth points the other way: whatever inflates heterozygosity in
these samples is concentrated in the ones that sequenced worst, which is the
pattern a mixed or contaminated sample produces. Either way a comparison between
groups that differ in depth is confounded.

```{r}
het_per_sample |>
  ggplot(aes(x = mean_dp, y = pct_het, color = species)) +
  geom_point(size = 2.5, alpha = 0.8) +
  labs(
    title = "Depth vs. heterozygosity",
    x = "Mean depth of coverage",
    y = "Percentage of called sites that are heterozygous",
    color = "Species"
  ) +
  theme(legend.text = element_text(face = "italic"))
```

### Heterozygosity and sequencing batch

Within *B. jararaca*, heterozygosity and singleton counts are higher in the 605
batch, which is also deeper, so this has the shape of an allele-dropout artifact.
It is not one: within *jararaca*, depth barely correlates with heterozygosity,
and the difference follows sampling locality, which is confounded with batch
(RS, RJ and PR are almost entirely 705; SC and ES mostly 605). Adding batch on
top of `state` explains nothing further, while depth and missingness keep a
strong batch effect after conditioning on state. The *state* comparison in 05c
therefore needs the most care.

```{r}
het_per_sample |>
  filter(species == "jararaca") |>
  summarize(
    n = n(),
    pct_het = mean(pct_het),
    mean_dp = mean(mean_dp),
    n_singletons = mean(n_singletons),
    .by = c(state, seq_batch)
  ) |>
  arrange(state, seq_batch) |>
  qc_gt(digits = 3, title = "B. jararaca by state and batch", full_width = TRUE)
```

Sex calls and locus filters are not affected: the Z:autosome depth ratio is a
within-sample ratio and matches between batches, and the per-locus ordering of
heterozygosity and missingness is preserved across batches.

## Variant quality per sample

Real SNPs have a transition:transversion ratio of ~2, whereas random calls have 0.5.

```{r}
sample_final |>
  ggplot(aes(x = ts_tv)) +
  geom_histogram(binwidth = 0.01, fill = "steelblue", color = "black") +
  geom_vline(xintercept = 2, linetype = "dashed", color = "grey40") +
  labs(
    title = "Ts/Tv ratio per sample",
    x = "Transitions / transversions",
    y = "Frequency (samples)"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

# A ratio that falls away at low depth is the signature of depth-driven noise
sample_final |>
  ggplot(aes(x = mean_dp, y = ts_tv)) +
  geom_point(alpha = 0.7, size = 2.5, color = "steelblue") +
  geom_hline(yintercept = 2, linetype = "dashed", color = "grey40") +
  labs(
    title = "Depth vs. Ts/Tv",
    x = "Mean depth of coverage",
    y = "Transitions / transversions"
  )
```

Private alleles are the other signature of contamination, so a sample far above
its peers is worth a look even when its Ts/Tv is normal. Read them next to
missingness: a sample with more sites called than its peers has more opportunity
to carry a private allele, so the count is not purely a contamination signal.

```{r}
sample_final |>
  ggplot(aes(x = n_singletons)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black") +
  labs(
    title = "Singleton (private) variants per sample",
    x = "Number of singletons",
    y = "Frequency (samples)"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))
```

### The singleton outlier: I20937

```{r}
# Values for the text below and in 'Samples to watch'
sv <- function(s, col) sample_final[[col]][sample_final$sample == s]
b705 <- filter(sample_final, seq_batch == "705_H2HLLBCXY")
```

I20937 has `r format(sv("I20937", "n_singletons"), big.mark = ",")` singletons
against a median of `r median(sample_final$n_singletons)` ---
`r sprintf("%.1f%%", 100 * sv("I20937", "n_singletons") / sum(sample_final$n_singletons))`
of every singleton site in the data set, from one of `r nrow(sample_final)`
samples. It is the only *B. jabrensis* in the panel, so every difference from the
other species appears as a singleton. Being the only sample of its species is not
enough on its own: *B. germanoi* (I1773) is also n = 1 but has
`r sv("I1773", "n_singletons")` singletons, so divergence from *B. jararaca*, on
which the probes were designed, matters.

```{r}
sample_final |>
  summarize(
    n_samples = n(),
    median_singletons = median(n_singletons),
    median_pct_on_target = median(pct_on_target),
    median_enrichment = median(enrichment),
    .by = species
  ) |>
  arrange(desc(median_singletons)) |>
  qc_gt()
```

It does not look contaminated: heterozygosity is among the lowest in the panel
(`r sprintf("%.2f%%", het_per_sample$pct_het[het_per_sample$sample == "I20937"])`
vs a median of `r sprintf("%.2f%%", median(het_per_sample$pct_het))`) and Ts/Tv is
normal (`r round(sv("I20937", "ts_tv"), 2)`). Capture is typical for its batch
(705: `r round(sv("I20937", "pct_on_target"), 1)`% on target, rank
`r rank(b705$pct_on_target)[b705$sample == "I20937"]` of `r nrow(b705)`), but its
error rate (`r signif(sv("I20937", "error_rate"), 3)`) and missingness
(`r sprintf("%.1f%%", sv("I20937", "mean_pct_N_final"))`) are the highest in the
panel, consistent with reads diverging from the reference. Usable, but expect it
to look artificially diverged and homozygous against the *jararaca* bulk.

## Samples to watch

Every section above flags its own outliers. This pulls them into one list, so
that a sample failing on several counts is visible as such rather than appearing
once in each table. The thresholds are deliberately loose --- this is a list to
look at, not a filter.

The quantiles are taken **within** each sequencing batch. Panel-wide thresholds
cut the tail of a bimodal distribution instead of finding outliers, so they
mostly selected a batch: of the seven samples flagged for low depth all seven
were 705, of the seven flagged for high duplication all seven were 605, and the
singleton and Ts/Tv flags went 6 of 7 to 605. A within-batch quantile asks the
question that was meant --- whether a sample is unusual next to the samples it
was sequenced with.

```{r}
flags <- sample_final |>
  left_join(select(ab_sample, sample, pct_blanked, median_ab), by = "sample") |>
  mutate(
    `low depth` = mean_dp < quantile(mean_dp, 0.05),
    `high missingness` = mean_pct_N_final > quantile(mean_pct_N_final, 0.95),
    `low Ts/Tv` = ts_tv < quantile(ts_tv, 0.05),
    `many singletons` = n_singletons > quantile(n_singletons, 0.95),
    `high duplication` = pct_duplication > quantile(pct_duplication, 0.95),
    `low mapping` = pct_mapped < quantile(pct_mapped, 0.05),
    # From the allele-balance section: many hets outside the 02b window, and a
    # balance skewed toward the reference, the signature of a second DNA source
    `many hets blanked` = pct_blanked > quantile(pct_blanked, 0.95),
    `low allele balance` = median_ab < quantile(median_ab, 0.05),
    .by = seq_batch
  ) |>
  # Not batch-relative: the confidence flag comes from 04b, which already
  # z-scores within the assigned sex group
  mutate(`sex not confident` = !sex_confident) |>
  mutate(n_flags = rowSums(pick(`low depth`:`sex not confident`)))

watch <- flags |>
  filter(n_flags >= 2) |>
  arrange(desc(n_flags))

# Low allele balance against sex-call confidence, for the text below
ab_sex <- with(flags, table(
  low_ab = factor(`low allele balance`, levels = c(FALSE, TRUE)),
  not_confident = factor(`sex not confident`, levels = c(FALSE, TRUE))
))
ab_sex_p <- fisher.test(ab_sex)$p.value

watch |>
  select(sample, species, seq_batch, n_flags, mean_dp, mean_pct_N_final, ts_tv,
         n_singletons, pct_duplication, pct_mapped, pct_blanked, median_ab) |>
  qc_gt()
```

Which checks each listed sample failed:

```{r}
watch |>
  select(sample, n_flags, `low depth`:`sex not confident`) |>
  pivot_longer(!c(sample, n_flags), names_to = "check") |>
  filter(value) |>
  summarize(failed = paste(check, collapse = ", "), .by = c(sample, n_flags)) |>
  arrange(desc(n_flags)) |>
  qc_gt(full_width = TRUE)
```

The two allele-balance flags and the sex-calling flag keep landing on the same
samples, and they are measured independently: allele balance on the autosomal
loci, sex from Z depth and Z heterozygosity. Of the `r sum(ab_sex["TRUE", ])`
samples with low median allele balance for their batch,
`r ab_sex["TRUE", "TRUE"]` are also 'sex not confident', against
`r ab_sex["FALSE", "TRUE"]` of the other `r sum(ab_sex["FALSE", ])` (Fisher's
exact test, p = `r format(signif(ab_sex_p, 1), scientific = TRUE)`). How they fail is consistent with a
second source of DNA:

- **I1495 and I1507** are females whose Z depth ratio is well above the female
  norm (`r round(sv("I1495", "z_depth_ratio"), 2)` and
  `r round(sv("I1507", "z_depth_ratio"), 2)`, against a female median of
  `r round(median(sample_final$z_depth_ratio[sample_final$sex == "female"]), 2)`),
  which is what reads from a male would add.
- **I1624** is a male flagged on both its depth ratio and its heterozygous
  gametolog sites.
- **I1671** fits the same pattern more weakly: a female with a raised depth ratio
  (`r round(sv("I1671", "z_depth_ratio"), 2)`) and low balance for its batch.
- Among females, the three with the lowest allele balance --- I1495, I1507, and
  I1520 --- have three of the four highest Z het densities, all above the 95th
  percentile for females. A ZW female should have essentially no Z
  heterozygosity.

None of these samples is among those flagged as contaminated in the metadata.
I1495, I1507, and I1624, and I1517 on allele balance alone, are the ones to
check first. The three samples with the most hets blanked (I1768, I1670, I1684)
are not low in allele balance and have confident sex calls, and their balance is
centered on 0.5, so they do not share this pattern.

## Metadata-flagged samples

The `r n_contam` samples flagged as contaminated in the metadata were removed in
03c, before any locus filter, and their stats come from side runs (step 4I).
Singleton counts and sex-call confidence depend on which samples are scored
together, so in this section both groups take them from the 4I runs, which
scored all `r nrow(vcf_stats_flagged)` samples together; elsewhere in this report
the retained samples' values come from runs without the flagged samples. A
second DNA source should show as a lower median allele balance, more hets below
0.35, and more sex calls that are not confident.

```{r}
# Heterozygosity is 05a's pooled `het_pct_final` here, since `all_final` only
# holds the retained samples
flagged_cmp <- per_sample |>
  filter(retained | contam %in% TRUE) |>
  mutate(group = ifelse(retained, "retained", "flagged")) |>
  select(!c(n_singletons, sex_confident)) |>
  left_join(select(vcf_stats_flagged, sample, n_singletons), by = "sample") |>
  left_join(transmute(sex_flagged, sample, sex_confident = confident == "yes"),
            by = "sample")

flagged_cmp |>
  summarize(
    n = n(),
    across(c(mean_dp, mean_pct_N_final, het_pct_final, ab_median,
             ab_pct_below_0.35, ab_pct_blanked, n_singletons, ts_tv),
           \(x) median(x, na.rm = TRUE)),
    pct_sex_not_confident = 100 * mean(!sex_confident, na.rm = TRUE),
    .by = group
  ) |>
  qc_gt(title = "Medians, flagged vs. retained samples", full_width = TRUE)

flagged_cmp |>
  ggplot(aes(x = ab_median, y = het_pct_final, color = group)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Allele balance vs. heterozygosity",
    x = "Median allele balance of evaluated hets",
    y = "Percentage of called sites that are heterozygous",
    color = NULL
  )
```

`r n_low_ab_flagged` of the `r n_contam` flagged samples have a median allele
balance below `r ab_low`, against `r n_low_ab_retained` of the
`r nrow(sample_final)` retained samples. The flagged samples with low balance or
a sex call that is not confident are below. Most of the second kind combine a
balance of exactly 0.5 with very few hets and a lot of missing data. Across all
flagged samples,
`r sum(flagged_cmp$group == "flagged" & flagged_cmp$mean_pct_N_final > 10)` of
`r n_contam` have more than 10% missing data over the final loci, so the
missingness filter would likely have removed them anyway.

```{r}
flagged_cmp |>
  filter(group == "flagged", ab_median < ab_low | !sex_confident) |>
  select(sample, species, seq_batch, mean_dp, mean_pct_N_final, het_pct_final,
         ab_median, ab_pct_below_0.35, n_singletons, sex_confident) |>
  arrange(ab_median) |>
  qc_gt(digits = 3)
```

<details><summary>R session info</summary>

```{r}
#| echo: true
sessionInfo()
```

</details>
