---
title: "Explore probe sequence BLAST results"
author: "Jelmer Poelstra"
date: today
format:
  html:
    self-contained: true
    cache: false
    number-sections: true
    number-depth: 2
    toc: true
    fig-dpi: 300
    df-print: kable
    grid:
      body-width: 800px
knitr:
  opts_chunk:
    message: FALSE
    out.width: "85%"
    R.options: 
      knitr.table.html.attr: "quarto-disable-processing=true"
execute: 
  warning: false
editor_options: 
  chunk_output_type: console
---

<br>

-------

```{r}
# Load packages
library(tidyverse)
library(here)

# Set plotting theme
theme_set(theme_bw(base_size = 14))
theme_update(panel.grid.minor = element_blank())

# Define input files
blast_file <- here("results/blast_probes/blast_out_final.tsv") 
probe_lens_file <- here("results/blast_probes/probe_lengths.txt")

# Define output files
locus_bed_file <- here("results/blast_probes/probes.bed")
```

```{r}
# Read input files
blast <- read_tsv(blast_file, show_col_types = FALSE) |>
  select(
    q = qseqid, s = sacc, pident, len = length, qlen, qcov = qcovhsp,
    chrom = sacc, start = sstart, end = send
  )

lens <- read_tsv(probe_lens_file, col_names = c("probe", "len"))
```

<br>

## Best hit for each probe

```{r}
best_hits <- blast |> group_by(q) |> slice_head(n = 1)

best_hits |>
  ggplot(aes(x = pident, y = qcov)) +
  geom_jitter(
    shape = 21,
    size = 2,
    fill = "steelblue",
    alpha = 0.5,
    width = 0,
    height = 0.25
    ) +
  scale_x_continuous(
    labels = scales::label_percent(scale = 1),
    expand = expansion(mult = c(0.01, 0.02))
    ) +
  scale_y_continuous(
    labels = scales::label_percent(scale = 1),
    expand = expansion(mult = c(0.01, 0.02))
    ) +
  labs(x = "Identity", y = "Query coverage")
```

- Number of probes with a hit (identity >= 90, coverage >= 80%):

```{r}
nrow(best_hits)
```

- Of these, number with a best hit with identity < 95%:

```{r}
best_hits |> filter(pident < 95) |> nrow()  # 18
```

- Of these, number with a best hit with query coverage < 90%:

```{r}
best_hits |> filter(qcov < 90) |> nrow()    # 16
```

- Remove hits with identity < 95% and/or coverage < 90%:

```{r}
blast_filt <- blast |>
  group_by(q) |>
  filter(any(pident >= 95)) |>
  filter(any(qcov >= 90)) |>
  ungroup()
```

<br>

## Multi-hits

```{r}
multi_hits <- blast_filt |>
  group_by(q) |>
  add_count(name = "n_hits") |>
  filter(n_hits > 1) |>
  mutate(
    id_drop = pident - lead(pident),
    cov_drop = qcov - lead(qcov),
  )
```

- Number of loci with multiple hits:

```{r}
length(unique(multi_hits$q)) # 83
```

```{r}
#- Keep only those with a drop in percent ID of at least 5,
#  and with percent ID of at least 95 and percent coverage of at least 90:

#multi_hits_keep <- multi_hits |>
#  slice_head(n = 1) |>
#  filter(id_drop >= 5, pident >= 95, qcov >= 90) |>
#  pull(q)

#mutate(status = ifelse(q %in% multi_hits_keep, "keep", "remove")) |>
```

- For multi-hit probes, plot showing percent ID for each hit:

```{r}
multi_hits |>
  ungroup() |>
  arrange(desc(pident)) |>
  mutate(q = fct_inorder(q)) |>
  ggplot(aes(x = q, y = pident)) +
  geom_line(color = "grey30") +
  geom_point(shape = 21, size = 2.5, fill = "red") +
  scale_y_continuous(
    labels = scales::label_percent(scale = 1),
    expand = expansion(mult = c(0.01, 0.02))
    ) +
  labs(x = "Query", y = "Percent identity") +
  theme(
    axis.text.x = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.position = "top"
  )
```

- Remove multi-hit probes:

```{r}
blast_filt2 <- blast_filt |>
  filter(!q %in% unique(multi_hits$q)) |>
  group_by(q) |>
  slice_head(n = 1) |>
  ungroup()
```

- Create BED files:

```{r}
locus_bed <- blast_filt2 |>
  mutate(
    start_final = ifelse(start < end, start, end) - 1,
    end_final = ifelse(start < end, end, start)
  ) |>
  select(chrom, start = start_final, end = end_final, probe = q)
  #mutate(start = start - 1, score = 0, strand = ".") |>
  #select(chrom, start, end, probe = q, score, strand)

#locus_bed_buffer <- locus_bed |>
  
write_tsv(locus_bed, locus_bed_file, col_names = FALSE)
```

- Number of loci left:

```{r}
nrow(blast_filt2)
```

<br>

## Probe lengths

```{r}
ggplot(lens, aes(x = len)) +
  geom_histogram() +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.02)),
    labels = scales::comma
    ) +
  scale_x_continuous(labels = scales::comma) +
  labs(x = "Probe length (bp)")
```
